---
title: "Fig-S9: T4 & T3 RT-qPCR"
author: "Andrea Mariossi"
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: false
    code-line-numbers: true
    page-layout: article
execute:
  echo: true
  warning: false
  message: true
fig-width: 12
fig-height: 8
fig-align: center
editor_options:
  chunk_output_type: inline
---

## Setup

Inputs come either from my local folder or from the repository's folder (when downloaded from GitHub).

Outputs are written to the `output/` folder.

```{r setup}
rm(list = ls())

# ---- packages ----
suppressPackageStartupMessages({
  library(here)
  library(tidyverse)
  library(readxl)
  library(paletteer)
  library(rstatix)
  library(ggpubr)
  library(knitr)
  library(Cairo)
  library(lme4)
  library(lmerTest)
  library(emmeans)
  library(kableExtra)
  library(patchwork)
})
```

## Configuration

```{r config}
# ---- configuration ----

pick_pipeline <- "local"  # "local" or "github"

# ---- my paths (inputs+outputs) ----
file_path <- "/Users/andrea/Library/CloudStorage/GoogleDrive-andrea.mariossi@gmail.com/My Drive/4.paper_archive/2025_thyroid_pathway_&_metamorphosis/04.SciAdv_rev/source_data"
out_dir   <- file.path(file_path, "output")

# ---- github repo paths ----
repo_dir <- here()

# fallback: check if repo exists in Downloads folder
if (!dir.exists(file.path(repo_dir, "data"))) {
  repo_dir_try <- path.expand("~/Downloads/thyroid_hormone_ciona-main")
  if (dir.exists(file.path(repo_dir_try, "data"))) {
    repo_dir <- repo_dir_try
  }
}

# ---- switch ----
if (pick_pipeline == "local") {
  data_dir <- file_path
  SAVE_OUTPUTS <- TRUE
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
  opath <- function(...) file.path(out_dir, ...)
  
} else if (pick_pipeline == "github") {
  data_dir <- file.path(repo_dir, "data")
  stopifnot(dir.exists(data_dir))
  
  SAVE_OUTPUTS <- TRUE
  out_dir <- file.path(repo_dir, "output")
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
  opath <- function(...) file.path(out_dir, ...)
  
} else {
  stop("pick_pipeline must be either 'local' or 'github'")
}

dpath <- function(...) file.path(data_dir, ...)

message("Pipeline: ", pick_pipeline)
message("Data dir (inputs): ", data_dir)
message("Outputs enabled: ", SAVE_OUTPUTS)
if (SAVE_OUTPUTS) message("Outputs dir: ", out_dir)
```

## Theme & Style

```{r theme}
# ---- theme & style ----
custom_theme <- function() {
  theme(
    axis.ticks = element_line(linewidth = 0.5),
    axis.ticks.length = unit(0.09, "cm"),
    legend.key.height = unit(20, "pt"),
    axis.line = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank(),
    strip.background = element_blank()
  )
}

# ===== theme & style =====
theme_set(theme_pubr(base_size = 13, base_family = "sans") + custom_theme())

# ---- color palette ----
contPalette <- paletteer_c("ggthemes::Red-Green-Gold Diverging", 30)

# ===== extra bits =====
no_legend <- theme(legend.position = "none")
axis_text_45 <- theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
stage_order <- c("earTII", "midTII", "latTII", "latTIII", "larva")


# ---- pretty p-values for kable ----
format_pvalues <- function(df) {
  df %>%
    mutate(across(
      any_of(c("p.value", "p.value.x", "p.value.y", "adj.p.value")),
      ~ case_when(
        . < 0.001 ~ format(., scientific = TRUE, digits = 2),
        TRUE ~ as.character(round(., 4))
      )
    ))
}
```

## Load Data

```{r load_data}
# -------------------------------------
# qPCR +/- T4 & TT3
# -------------------------------------

qpcr_file <- dpath("7_T4_T3_qPCR.xlsx")
stopifnot(file.exists(qpcr_file))

# read T4
qPCR_T4 <- read_excel(qpcr_file, range = cell_cols("A:I")) %>%
  mutate(across(where(is.character), tolower))

# read T3
qPCR_T3 <- read_excel(qpcr_file, range = cell_cols("K:S")) %>%
  mutate(across(where(is.character), tolower))

kable(head(qPCR_T4), caption = "T4 Treatment Data (first 6 rows)")
kable(head(qPCR_T3), caption = "T3 Treatment Data (first 6 rows)")
```

## Quality Control: Raw Ct Values

```{r qc_raw_ct}

# ---- average Ct values for technical replicates ----

process_one_qpcr_data <- function(data) {
  # calculating average Ct values for technical replicates
  summarised_data <- data %>%
    mutate(gene = as.factor(gene)) %>%
    rowwise() %>%
    mutate(mean_Ct = (ct_1 + ct_2) / 2) %>%
    ungroup()
  
  # plot average Ct values
  ct_plot <- ggplot(summarised_data, aes(x = group, y = mean_Ct, colour = hpf)) +
    geom_point() +
    scale_color_gradientn(colors = contPalette, name = "early to late hpf") +
    facet_wrap(~ gene, scales = "free") +
    custom_theme()
  
  # print(ct_plot)
  
  list(data = summarised_data, plot = ct_plot)
}

process_one_qpcr_data(qPCR_T4) -> Ct_T4 
Ct_T4$plot + ggtitle("T4") + Ct_T4$plot+ facet_wrap(~stage)

process_one_qpcr_data(qPCR_T3)  -> Ct_T3
Ct_T3$plot + ggtitle("T3") + Ct_T3$plot + facet_wrap(~stage)
```

### Confirm Actin Stability

```{r actin_stability}
# ---- confirm actin Ct values don't change with treatment/development ----

run_actin_qc <- function(input_df, label = "T4 / T3") {
  df <- input_df$data
  
  anova_rawCt_tbl <- df %>%
    group_by(gene) %>%
    nest() %>%
    mutate(
      model = map(data, ~ aov(mean_Ct ~ group * stage, data = .x)),
      tidied = map(model, broom::tidy)
    ) %>%
    unnest(tidied) %>%
    select(gene, term, df, sumsq, meansq, statistic, p.value) %>%
    mutate(
      p.value = round(p.value, 6),
      significance = case_when(
        p.value < 0.001 ~ "***",
        p.value < 0.01 ~ "**",
        p.value < 0.05 ~ "*",
        TRUE ~ "ns"
      )
    )
  
  actin_plot <- ggplot(
    df %>% filter(gene == "actin"),
    aes(x = group, y = mean_Ct, colour = hpf)
  ) +
    geom_point(size = 2, alpha = 0.7) +
    scale_color_gradientn(colors = contPalette, name = "early to late hpf") +
    facet_wrap(~ gene, scales = "free_y") +
    labs(title = paste0("Actin Ct Values (", label, ")")) +
    custom_theme()
  
  # print(actin_plot)
  pretty_table <- anova_rawCt_tbl |> format_pvalues() %>%
    kable(
      digits = 4,
      caption = paste0("Quality Control: ANOVA raw Ct showing actin stability (", label, ")")
    ) %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
  list(anova = pretty_table, plot = actin_plot)
}

# ---- T4 actin summary ----
run_actin_qc(Ct_T4, label = "T4") -> qc_T4_actin
qc_T4_actin$plot + ggtitle("T4")
qc_T4_actin$anova

# ---- T3 actin summary ----
run_actin_qc(Ct_T3, label = "T3")  -> qc_T3_actin
qc_T3_actin$plot + ggtitle("T3")
qc_T3_actin$anova 

```

Actin shows non-significant effects - reference gene is stable both in T4 and T3.

## ΔCt Calculation

```{r delta_ct}
# ---- ΔCt function ----

# delta-Ct method: difference between Ct of housekeeping gene and target gene
process_two_qpcr_data <- function(summarised_data) {
  
  # rearranging the data
  goi_data <- summarised_data %>%
    filter(gene != "actin") %>%
    rename("goiMean_Ct" = "mean_Ct")
  
  ref_data <- summarised_data %>%
    filter(gene == "actin") %>%
    rename("refMean_Ct" = "mean_Ct")
  
  # delta-Ct method: difference between Ct of HK gene and target gene
  # RNA relative 2^delta-Ct
  delta_Ct <- left_join(goi_data, ref_data, 
                        by = c("group", "hpf", "stage", "bioRep")) %>%
    rename(
      gene_goi = gene.x,
      goi_Ct = goiMean_Ct,
      ref_Ct = refMean_Ct
    ) %>%
    relocate(group, bioRep, everything()) %>%
    mutate(
      delta_Ct = goi_Ct - ref_Ct,  # delta-Ct method
      mRNArel = 2^(-delta_Ct)       # relative mRNA
    )
  
  # plot delta_Ct each sample
  delta_ct_plot <- ggplot(delta_Ct, aes(x = group, y = delta_Ct, colour = hpf)) +
    geom_point() +
    scale_color_gradientn(colors = contPalette, name = "early to late hpf") +
    facet_wrap(~ gene_goi, scales = "free") +
    labs(
      x = "",
      y = "ΔCt (Target Gene - Reference)"
    ) +
    custom_theme()
  
  print(delta_ct_plot)
  
  delta_Ct    
  # return(delta_Ct)
}

# process both datasets using the same function
delta_Ct_T4 <- process_two_qpcr_data(Ct_T4$data)
delta_Ct_T3 <- process_two_qpcr_data(Ct_T3$data)

kable(head(delta_Ct_T4), caption = "ΔCt Data for T4 (first 6 rows)")
```

## Quick Look at Expression Patterns

```{r quick_look, fig.width=16, fig.height=8}
# ---- plotting function ----

create_qpcr_plot <- function(input_data, treatment_color) {
  # x-axis breaks and labels
  x_breaks <- unique(input_data$hpf)
  x_labels <- c("17.3 \n hatch", "21.3 \n earSwim", "24.5 \n midSwim",
                "27.5 \n latSwim", "29 \n attach")
  
  # y-axis breaks and labels
  y_range <- range(input_data$mRNArel, na.rm = TRUE)
  y_breaks <- round(10^(seq(log10(y_range[1]), log10(y_range[2]), length.out = 5)), 4)
  
  ggplot(input_data, aes(
    x = hpf, y = mRNArel,
    group = group, color = group, fill = group, shape = as.factor(bioRep)
  )) +
    geom_point(alpha = 0.6, size = 3) +
    stat_summary(fun = mean, geom = "line", linewidth = 1.5) +
    stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.2, color = NA) +
    scale_x_continuous(breaks = x_breaks, labels = x_labels) +
    # scale_y_continuous(trans = "log10", breaks = y_breaks) +
    scale_color_manual(values = c("grey78", treatment_color)) +
    scale_fill_manual(values = c("grey78", treatment_color)) +
    labs(x = "Developmental stage (hpf)", y = "mRNA expression") +
    custom_theme() +
    facet_wrap(~gene_goi, ncol = 3, scales = "free_y") +
    theme(legend.position = "bottom")
}

# create plots
plot_t4 <- create_qpcr_plot(delta_Ct_T4, "#6baed6")  # control=grey, T4=blue
plot_t3 <- create_qpcr_plot(delta_Ct_T3, "#c9a8e0")  # control=grey, T3=purple

plot_t4
plot_t3
```

## Statistical Analysis

### Two-way ANOVA

Testing for:

-   **group**: Does T4 treatment have an overall effect on expression?
-   **stage**: Does expression vary by developmental stage?
-   **group:stage**: Does the effect of T4 change across stages?

The group:stage interaction tests whether the effect of the drug (T4/T3) on gene expression changes depending on the developmental stage.

-   if the effect of T4/T3 is consistent across all stages → this interaction will not be significant.
-   if the effect of T4/T3 is stronger at some stages and weaker or absent at others → the interaction will be significant.

```{r anova_tukey}
# ---- two-way ANOVA for each gene ----

run_anova_tukey <- function(delta_df) {
  delta_df %>%
    group_by(gene_goi) %>%
    nest() %>%
    mutate(
      # two-way ANOVA per gene
      anova_df = map(data, ~ aov(delta_Ct ~ group * stage, data = .x)),
      anova_tidy = map(anova_df, broom::tidy),
      
      # tukey post-hoc per gene (all terms)
      tukey_df = map(anova_df, TukeyHSD),
      tukey_tidy = map(tukey_df, broom::tidy)
    )
}

# anova is the first pass, they look ok so let directly move on to Tukey to see in which groups/stages is that effect?
# aov_results_T4 <- aov_tukey_T4 %>%
#   select(gene_goi, anova_tidy) %>%
#   unnest(anova_tidy)
# 
# aov_results_T3 <- aov_tukey_T3 %>%
#   select(gene_goi, anova_tidy) %>%
#   unnest(anova_tidy)

aov_tukey_T4 <- run_anova_tukey(delta_Ct_T4)
aov_tukey_T3 <- run_anova_tukey(delta_Ct_T3)

aov_tukey_results_T4 <- aov_tukey_T4 %>%
  select(gene_goi, tukey_tidy) %>%
  unnest(tukey_tidy)

aov_tukey_results_T3 <- aov_tukey_T3 %>%
  select(gene_goi, tukey_tidy) %>%
  unnest(tukey_tidy)

aov_tukey_results_T4 %>%
  select(-"null.value") %>%
  filter(term %in% c("group", "stage")) %>%  # no interaction term
  format_pvalues() %>%
  kable(caption = "T4: Tukey HSD Post-hoc Tests") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

aov_tukey_results_T3 %>%
  select(-"null.value") %>%
  filter(term %in% c("group", "stage")) %>%  # no interaction term
  format_pvalues() %>%
  kable(caption = "T3: Tukey HSD Post-hoc Tests") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r}
# -------------------------------------
# statistics on delta Ct 
# -------------------------------------

# this is to check each stage separatly ==> not needed
# # wilcox_test
# stat_results_T4 <- delta_Ct_T4 %>%
#   group_by(gene_goi, stage) %>%
#   wilcox_test(delta_Ct ~ group) %>%    # compare ΔCt between Control vs. T4
#   add_significance() %>%
#   adjust_pvalue(method = "fdr") %>%  # correct for multiple tests
#   add_significance()
# 
# # pretty  table
# stat_results_T4 %>%
#   dplyr::select(gene_goi, stage, 
#          p, p.adj, 
#          p.signif, p.adj.signif) %>%
#   knitr::kable(digits = 4)
# 
# stat_results_T3 <- delta_Ct_T3 |> 
#   group_by(gene_goi, stage) %>%
#   # adjust_pvalue(method = "bonferroni") %>%
#   wilcox_test(delta_Ct ~ group) %>%    # compare ΔCt between Control vs. T4
#   add_significance() %>%
#   adjust_pvalue(method = "fdr") %>%  # correct for multiple tests
#   add_significance()
# 
# # pretty  table
# stat_results_T3 %>%
#   dplyr::select(gene_goi, stage, 
#          p, p.adj, 
#          p.signif, p.adj.signif) %>%
#   knitr::kable(digits = 4)
```

## Main Figure: Panel Plot S9

```{r panel_plot, fig.width=18, fig.height=16}
# ---------------------------------------------------
# Plotting same y-axes ~gene & shared legend 
# ---------------------------------------------------

# ---- combine T4 and T3 data ----
combined_data <- bind_rows(
  delta_Ct_T4 %>% mutate(treatment = "T4"),
  delta_Ct_T3 %>% mutate(treatment = "T3")
)

# ---- per-gene global y-limits across T4 + T3 ----
gene_ylimits <- combined_data %>%
  group_by(gene_goi) %>%
  summarise(
    ymin = min(mRNArel, na.rm = TRUE),
    ymax = max(mRNArel, na.rm = TRUE),
    .groups = "drop"
  )

print(gene_ylimits)

# shared palette & order
pal_vals <- c("Control" = "grey78", "T4" = "#6baed6", "T3" = "#c9a8e0")
brks <- c("Control", "T4", "T3")

# ---- plot function ----
create_qpcr_plot <- function(input_data, treatment_label) {
  
  # current gene (one per loop)
  current_gene <- unique(input_data$gene_goi)
  
  # x-axis breaks and labels
  x_breaks <- unique(input_data$hpf)
  x_labels <- c("17.3 \n hatch", "21.3 \n earSwim", "24.5 \n midSwim",
                "27.5 \n latSwim", "29 \n attach")
  
  # gene-wise y-limits
  y_limits <- gene_ylimits %>%
    filter(gene_goi == current_gene) %>%
    select(ymin, ymax) %>%
    as.list()
  
  # new column so i can collapse the legend
  input_data <- input_data %>%
    mutate(group2 = if_else(group == "control", "Control", treatment_label))
  
  ggplot(input_data, aes(hpf, mRNArel, color = group2, fill = group2)) +
    geom_point(alpha = 0.6, size = 3, show.legend = TRUE) +
    stat_summary(fun = mean, geom = "line", linewidth = 1.2, show.legend = FALSE) +
      # <- position = position_jitter(width = 0.1, height = 0)
    stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.2, 
                 color = NA, show.legend = FALSE) +
    scale_x_continuous(breaks = x_breaks, labels = x_labels) +
    scale_color_manual(values = pal_vals, breaks = brks, limits = brks, name = NULL) +
    scale_fill_manual(values = pal_vals, breaks = brks, limits = brks, name = NULL) +
    coord_cartesian(ylim = c(y_limits$ymin, y_limits$ymax)) +
    labs(x = "Developmental stage (hpf)", y = "mRNA expression") +
    custom_theme() +
    theme(axis.title.x = element_blank()) +
    ggtitle(current_gene)
}

# ---- per-gene panels for T4 ----
genes <- union(delta_Ct_T4$gene_goi, delta_Ct_T3$gene_goi) %>% unique()
# genes <- unique(delta_Ct_T4$gene_goi) #  unique genes

plot_list <- map(genes, function(gene) {
  # filter + individual plot
  t4 <- delta_Ct_T4 %>% filter(gene_goi == gene) %>% create_qpcr_plot("T4")
  t3 <- delta_Ct_T3 %>% filter(gene_goi == gene) %>% create_qpcr_plot("T3")
  t4 / t3  # stack T4 over T3
})

# ---- arrange all genes in a grid and collect a single legend ----
final_plot <- wrap_plots(plot_list, ncol = 3) +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 25),
    plot.title = element_text(hjust = 0.5, size = 20, face = "bold")
  ) &
  guides(
    color = guide_legend(
      title = NULL,
      override.aes = list(shape = 22, size = 12),
      nrow = 1
    ),
    fill = guide_legend(title = NULL, nrow = 1)
  )

final_plot
```

## Save Plots

```{r save_plots}
if (SAVE_OUTPUTS) {
  
  width_in <- 40
  aspect_ratio <- 3/4
  height_in <- width_in * aspect_ratio
  
  width_mm <- 4 * width_in / 25.4
  height_mm <- 4 * height_in / 25.4
  
  # save with CairoPDF
  CairoPDF(
    opath("figS9_rtqPCR_T4&T3.pdf"),
    width = width_mm,
    height = height_mm
  )
  print(final_plot)
  dev.off()
  
  # save with cowplot (aspect ratio)
  cowplot::save_plot(
    plot = final_plot,
    base_height = height_in,
    base_asp = width_in / height_in,
    file = opath("figS9_rtqPCR_T4&T3_1.pdf")
  )
  
  cowplot::save_plot(
    plot = final_plot,
    base_height = 0.9 * 3.71,
    base_asp = 0.9 * 1.618,
    file = opath("figS9_rtqPCR_T4&T3_2.pdf")
  )
  
  cowplot::save_plot(
    plot = final_plot,
    base_height = 1.9 * 3.71,
    base_asp = 0.9 * 1.618,
    file = opath("figS9_rtqPCR_T4&T3_3.pdf")
  )
  
  message("\nPlots saved to: ", out_dir)
}
```

## Session Info

```{r session_info}
sessionInfo()
```
