---
title: "Fig-4D_rescueExp_photoreceptor"
author: "Andrea Mariossi"

format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    toc-depth: 5
    code-fold: false
    code-line-numbers: true
    page-layout: article
    grid:
      sidebar-width: 100px
      body-width: 2000px
      margin-width: 100px
      gutter-width: 3.5rem

execute:
  echo: true
  warning: false
  message: true

fig-width: 18
fig-height: 10
fig-align: center
fig-format: png
out-width: "90%"

editor_options:
  chunk_output_type: inline
---

# Photoreceptor Overexpression: Opsin1 & Opsin2 & Thr

For each biological replicate, embryos were dissected from a pooled parent set of 4–5 wild caught adults. Following fertilization, embryos were electroporated across four distinct experimental groups:

-   **Control**: Scrambled sgRNA.

-   **Knockdown**: sgRNA targeting *Thr*.

-   **Rescue (20µg)**: sgRNA targeting *Thr* co-electroporated with 20µg *Opsin2* plasmid.

-   **Rescue (40µg)**: sgRNA targeting *Thr* co-electroporated with 40µg *Opsin2* plasmid.

Larvae were cultured until reaching the settlement stage, at which point they were fixed. Attachment success was quantified as a binary outcome (Yes/No). The assay utilized a standardized "swirl test," where attachment was defined as the larva's ability to
remain adhered to the petri dish bottom while the water was swirled. Each petri dish contained approximately 40–60 embryos. The specific sample size per dish varied due to the volumetric transfer of embryos from the electroporation cuvettes via pipetting.
The entire experimental workflow was repeated across four independent biological runs on different days, with each condition represented in every run.

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

-   **Inputs** come either from my local folder or from the repository’s folder (when downloaded from GitHub).
-   **Outputs** of GitHub are written to a new local folder named `output_rerun/`

## Setup

```{r setup}
#| include: false

rm(list = ls())

# ---- packages ----
suppressPackageStartupMessages({
  library(here)
  library(tidyverse)
  library(readxl)
  library(paletteer)
  library(rstatix)
  library(ggpubr)
  library(knitr)
  library(Cairo)
  library(lme4)
  library(lmerTest)
  library(emmeans)
  library(kableExtra)
  library(patchwork)
  library(binom)
})
```

## Configuration

```{r config}
# ---- configuration ----
pick_pipeline <- "local"  # "local" or "github"

# ---- my paths (inputs+outputs) ----
file_path <- "/Users/andrea/Library/CloudStorage/GoogleDrive-andrea.mariossi@gmail.com/My Drive/4.paper_archive/2025_thyroid_pathway_&_metamorphosis/04.SciAdv_rev/source_data"
out_dir   <- file.path(file_path, "output_rerun")

# ---- github repo paths ----
repo_dir <- here()

# fallback: check if repo exists in Downloads folder
if (!dir.exists(file.path(repo_dir, "data"))) {
  repo_dir_try <- path.expand("~/Downloads/thyroid_hormone_ciona-main")
  if (dir.exists(file.path(repo_dir_try, "data"))) {
    repo_dir <- repo_dir_try
  }
}

# ---- switch ----
if (pick_pipeline == "local") {
  data_dir <- file_path
  SAVE_OUTPUTS <- TRUE
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
  opath <- function(...) file.path(out_dir, ...)
  
} else if (pick_pipeline == "github") {
  data_dir <- file.path(repo_dir, "data")
  stopifnot(dir.exists(data_dir))
  
  SAVE_OUTPUTS <- TRUE
  out_dir <- file.path(repo_dir, "output_rerun")
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
  opath <- function(...) file.path(out_dir, ...)
  
} else {
  stop("pick_pipeline must be either 'local' or 'github'")
}

dpath <- function(...) file.path(data_dir, ...)

message("Pipeline: ", pick_pipeline)
message("Data dir (inputs): ", data_dir)
message("Outputs enabled: ", SAVE_OUTPUTS)
if (SAVE_OUTPUTS) message("Outputs dir: ", out_dir)
```

## Theme & Style

```{r theme}
# ---- theme & style ----
custom_theme <- function() {
  theme(
    axis.ticks = element_line(linewidth = 0.5),
    axis.ticks.length = unit(0.09, "cm"),
    legend.key.height = unit(20, "pt"),
    axis.line = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank(),
    strip.background = element_blank()
  )
}

theme_set(theme_pubr(base_size = 13, base_family = "Helvetica") + custom_theme())

no_legend <- theme(legend.position = "none")
axis_text_45 <- theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

# ----extra ----
stage_order <- c("earTII", "midTII", "latTII", "latTIII", "larva")
panel_a_colors <- c("#636363", "#00AAAB", "#1970DB")

# ---- pretty p-values for kable ----
format_pvalues <- function(df) {
  df %>%
    mutate(across(
      any_of(c("p.value", "p.value.x", "p.value.y", "adj.p.value")),
      ~ case_when(
        . < 0.001 ~ format(., scientific = TRUE, digits = 2),
        TRUE ~ as.character(round(., 4))
      )
    ))
}
```

## Load Data

### Photoreceptors Thr KD & rescue with Opsin2 using Opsin1 enhancer

**Key Results:**

-   sgRNA_Thr shows reduced attachment compared to control
-   sgThr_opsin2_20ug shows increased attachment compared to control
-   40μg opsin2 cDNA shows even stronger effects

```{r load_data}
rescue_file <- dpath("13_photoreceptor_CRISPR_rescue.xlsx")
stopifnot(file.exists(rescue_file))

# read and preprocess data
opsn1_cas_rescue.df <- read_excel(rescue_file) %>%
  slice(1:17) |> 
  drop_na()

# 'lacz' as the reference level
opsn1_cas_rescue.df$condition <- factor(opsn1_cas_rescue.df$condition, 
                                        levels = c("sgRNA_scramble", "sgRNA_thr", "sgThr_opsin2_20ug"))

# QC number
sample_summary <- opsn1_cas_rescue.df %>%
  group_by(condition, bioRep) %>%
  summarise(n_larvae = sum(attached_larvae + swimming_larvae),
            n_attached = sum(attached_larvae),.groups = "drop") |> 
  group_by(condition) %>%
  summarise(
    n_replicates = n(),
    total_larvae = sum(n_larvae),
    min_larvae_per_rep = min(n_larvae),
    max_larvae_per_rep = max(n_larvae)
  )

sample_summary %>%
  kable(caption = "Per-condition summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# percentages per replicate
opsn1_cas_rescue.tidy <- opsn1_cas_rescue.df %>%
  mutate(total_larvae = swimming_larvae + attached_larvae) %>%
  pivot_longer(
    cols = c(attached_larvae, swimming_larvae),
    names_to = "status",
    values_to = "count"
  ) %>%
  group_by(condition, bioRep) %>%
  mutate(percent = round(count * 100 / sum(count), 1)) %>%  # % per replicate
  ungroup()

opsn1_cas_rescue.tidy %>%
  kable(caption = "Per-replicate percentages") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# summarize per condition
opsn1_cas_rescue_condition.tidy <- opsn1_cas_rescue.df %>%
  mutate(total_larvae = swimming_larvae + attached_larvae) %>%
  group_by(condition) %>%
  summarise(
    attached = sum(attached_larvae),
    not_attached = sum(swimming_larvae),
    total = sum(total_larvae)
  ) %>%
  pivot_longer(
    cols = c(attached, not_attached),
    names_to = "status",
    values_to = "count"
  ) %>%
  mutate(percent = count / total * 100)

opsn1_cas_rescue_condition.tidy %>%
  kable(caption = "Per-condition summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# summary data: mean and SE for each segment
opsn1_cas_rescue_summary.df <- opsn1_cas_rescue.tidy %>%
  filter(condition != "sgThr_opsin2_40ug") %>%
  group_by(condition, status) %>%
  summarise(
    mean_percent = mean(percent),
    se = sd(percent) / sqrt(n()),
    .groups = "drop"
  ) %>%
  arrange(condition, status) %>%
  # calculate cumulative y-position for error bars
  group_by(condition) %>%
  arrange(condition, desc(status)) %>%  # stacking order
  mutate(
    y_cumulative = cumsum(mean_percent),  # top of each segment
    y_start = y_cumulative - mean_percent  # bottom of each segment
  )

opsn1_cas_rescue_summary.df %>%
  kable(caption = "Summary statistics with error bar positions") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Quick Sanity Check

```{r quick_check}
# quick visualization
ggbarplot(
  opsn1_cas_rescue.tidy %>% drop_na(),
  x = "condition",
  y = "percent",
  add = "mean_se",
  error.plot = "errorbar",
  ylab = "% larvae attached",
  xlab = "",
  color = "black",
  fill = "status",
  palette = "uchicago",
  label = TRUE,
  lab.col = "black",
  lab.size = 5,
  lab.pos = "in",
  lab.nb.digit = 1
) +
  custom_theme()

# side-by-side comparison
opsn1_cas_rescue_summary.df %>%
  filter(condition != "sgThr_opsin2_40ug") %>%
  mutate(status = factor(status, levels = c("swimming_larvae", "attached_larvae"))) %>%
  ggplot(aes(x = condition, y = mean_percent, fill = status)) +
  geom_col(position = "dodge", width = 0.7, alpha = 0.8) +
  geom_errorbar(
    aes(ymin = mean_percent - se, ymax = mean_percent + se),
    position = position_dodge(width = 0.7),
    width = 0.2
  ) +
  scale_fill_manual(
    values = c("attached_larvae" = "#882222", "swimming_larvae" = "gray40"),
    labels = c("no", "yes")
  ) +
  scale_y_continuous(
    labels = scales::percent_format(scale = 1),
    breaks = c(25, 50, 75, 100),
    limits = c(0, 100)
  ) +
  labs(y = "% of Larvae", x = "Condition") +
  custom_theme()
```

## Stacked Plot with Symmetrical CI

```{r stacked_plot_symmetric}
#-------------------------------------------------------------------------------
# stacked plot
rescue_plot_onlyLower <- ggplot(
  opsn1_cas_rescue_summary.df,
  aes(x = condition, y = mean_percent, fill = status)
) +
  # stacked bars (mean %)
  geom_col(width = 0.7, alpha = 0.8, color = "black") +
  # segment-specific error bars
  geom_errorbar(
    data = subset(opsn1_cas_rescue_summary.df, status == "attached_larvae"),
    aes(
      ymin = y_start - se,  # at the boundary
      ymax = y_start + se
    ),
    width = 0.2,
    color = "black",
    linewidth = 0.5,
    position = position_dodge(width = 0)
  ) +
  geom_text(
    data = opsn1_cas_rescue_condition.tidy %>%
      filter(condition != "sgThr_opsin2_40ug") %>%
      distinct(condition, total),
    aes(x = condition, y = 105, label = paste0("n = ", total)),
    inherit.aes = FALSE
  ) +
  scale_fill_manual(
    values = c("attached_larvae" = "#882222", "swimming_larvae" = "gray70"),
    labels = c("Attached", "Not Attached")
  ) +
  labs(x = "Condition", y = "% larvae attached", fill = "Status") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top", panel.grid.major.x = element_blank())

rescue_plot_onlyLower +
  scale_y_continuous(
    breaks = seq(0, 100, by = 25),
    # labels = scales::label_number(suffix = "%"),
    expand = expansion(mult = c(0, 0.1))
  ) +
  # significance stars
  geom_text(
    data = data.frame(
      condition = c("sgRNA_thr", "sgThr_opsin2_20ug"),
      y = c(125, 125),
      label = c("**", "***") # this was *** wrong!
    ),
    aes(x = condition, y = y, label = label),
    inherit.aes = FALSE, size = 6, vjust = 0.5
  ) +
  # brackets for comparisons
  geom_segment(
    data = data.frame(
      x = c(1, 1),      # starting bar index
      xend = c(2, 3),   # ending bar index
      y = c(115, 120),  # height of the bracket
      yend = c(115, 120)
    ),
    aes(x = x, xend = xend, y = y, yend = yend),
    inherit.aes = FALSE
  ) +
  custom_theme()

rescue_plot_onlyLower + theme(
    legend.position = "bottom",
    panel.spacing.x = unit(0.1, "cm"),  # Tighter than 0.2
    axis.text.x = element_text(margin = margin(t = 2)),  # Less padding
    aspect.ratio = 1.2  
  )
```

## Batch consistency

```{r}
# ---- batch inspection  ----

# Calculate proportions per replicate and condition
batch_data <- opsn1_cas_rescue.tidy %>%
  filter(status == "attached_larvae" & condition != "sgThr_opsin2_40ug")

batch_plot <- ggplot(batch_data, aes(x = condition, y = percent, fill = condition)) +
  # Add a background area to show the "trend" per batch
  geom_line(aes(group = bioRep), alpha = 0.3, color = "gray50") +
  geom_point(shape = 21, size = 3, color = "black", stroke = 1) +
  # Facet by the biological replicate 
  facet_wrap(~bioRep, ncol = 4) + 
  scale_fill_manual(values = panel_a_colors) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    subtitle = "Each panel represents a different electroporation day",
    y = "% Larvae Attached",
    x = ""
  ) +
  custom_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  no_legend

batch_plot
```

## Statistical Analysis

Binomial outcomes (successes vs. failures).

Just to have an idea BUT IT IS NOT accurate, it doesn't consider biological replicate (pooling them)

-   Fisher's exact test: better for small samples
-   Chi-square test: large sample size

```{r statistical_tests}

#-------------------------------------------------------------------------------
# Fisher's Exact Test | Chi test
#-------------------------------------------------------------------------------

# ---- fisher's exact test | chi test ----
contingency_table <- opsn1_cas_rescue.df %>%
  drop_na() %>%
  mutate(total_larvae = swimming_larvae + attached_larvae) %>%
  group_by(condition) %>%
  summarise(
    attached = sum(attached_larvae),
    swimming = sum(swimming_larvae)
  ) %>%
  column_to_rownames("condition") %>%
  as.matrix()

### stat Fisher's Exact Test ==> 2.142e-14 p value
# fisher.test(contingency_table) # failing large sampel size ==> Chi-square Test	

# pearson's chi-square test
chi_test <- chisq.test(contingency_table)
chi_test

# perform all pairwise comparisons
pairwise_results <- pairwise.prop.test(
  contingency_table,
  p.adjust.method = "bonferroni"
) %>%
  tidy()

pairwise_results %>%
  kable(caption = "Pairwise comparisons (Bonferroni-corrected)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Logistic Regression

I used a **Generalized Linear Mixed Model (GLMM)** with **`bioRep` as a random intercept** to account for biological and technical variability inherent in **mosaic electroporation** and **wild-caught populations**.

-   The simplest model is a glm, but there is some variability in the electgroporation, the glmm wirh batch of larvae as a random effect works better to account for the variance.

-   **Random Effects**: The `bioRep` works as a **random intercept** to capture baseline shifts in attachment probability between parental sets and electroporation batches. The random effect `(1 | bioRep)` captures variability between biological replicates
    that the GLM ignores.

<!-- -->

-   **Fixed Effects**: Treatment conditions (Control, *Thr* Knockdown, and *Opsin2* Rescues) were treated as fixed effects.

-   Glmm is penalizing for additional parameter but still lower AIC (97.2 vs 107.1) and BIC (99.2 vs 108.6)

```{r logistic_regression}
# ---- logistic regression (for modeling probability of attachment) ----

# Standardizing the data object for all model comparisons
analysis_data <- opsn1_cas_rescue.df %>% 
  filter(condition != "sgThr_opsin2_40ug") %>%
  mutate(condition = factor(condition, 
                            levels = c("sgRNA_scramble", "sgRNA_thr", "sgThr_opsin2_20ug")))

#-----------------------------------------------------------------------------
# fit the simpler GLM (for comparison)
#-------------------------------------------------------------------------------
# ignores the batch/day effect
rescue_glm <- glm(
  cbind(attached_larvae, swimming_larvae) ~ condition,
  family = binomial,
  data = analysis_data
)

summary(rescue_glm)
# overdispersion check
performance::check_overdispersion(rescue_glm) # Significant overdispersion present

rescue_glm %>%
  tidy(conf.int = TRUE, exponentiate = TRUE) %>%
  kable(caption = "GLM Results (Odds Ratios)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# interpretation:
# - Intercept: Baseline odds (control = lacz)
# - sgRNA_thr: OR = 0.496  (50% lower odds vs control)
# - opsin2_20ug: OR = 2.63 ( 163% higher odds vs control)

#-----------------------------------------------------------------------------
# fit the robust GLMM (accounting for electroporation and shipping batches)
#-------------------------------------------------------------------------------
# mixed model: with random intercepts for replicates

rescue_glmer <- glmer(
  cbind(attached_larvae, swimming_larvae) ~ condition + (1 | bioRep),
  family = binomial,
  data = analysis_data
)


summary(rescue_glmer)
exp(fixef(rescue_glmer))  # get odds ratios
summary(rescue_glmer)$varcor # checking the "Random Effect" Variance
# extract confidence intervals for odds ratios
confint_model <- confint(rescue_glmer, method = "Wald")
confint_model

# extract p-values for the plot
# compare everything back to the intercept (Control)
stats_for_plot <- summary(rescue_glmer)$coefficients %>%
  as.data.frame() %>%
  rownames_to_column("group2") %>%
  filter(group2 != "(Intercept)") %>%
  mutate(group1 = "sgRNA_scramble",
          group2 = str_remove(group2, "condition"),
          p.adj = `Pr(>|z|)`, # You can apply p.adjust() here if desired
          label = case_when(p.adj < 0.001 ~ "***", p.adj < 0.01  ~ "**", p.adj < 0.05  ~ "*", TRUE ~ "ns")
         )
stats_for_plot

#-----------------------------------------------------------------------------
# ---- Model Comparison ----
# Likelihood Ratio Test to justify GLMM
lrt_result <- anova(rescue_glmer, rescue_glm)
lrt_result # pvale < 0.05 confirms GLMM is required

# overdispersion check
dispersion_ratio <- sum(residuals(rescue_glmer, type = "pearson")^2) / df.residual(rescue_glmer)
message("Dispersion Ratio: ", round(dispersion_ratio, 3))

performance::check_overdispersion(rescue_glmer)
# No significant overdispersion

#-----------------------------------------------------------------------------
# tidy results  
bind_rows(
  broom.mixed::tidy(rescue_glm, effects = "fixed", conf.int = TRUE) %>% mutate(model = "GLM"),
  broom.mixed::tidy(rescue_glmer, effects = "fixed", conf.int = TRUE) %>% mutate(model = "GLMM")
) %>%
  mutate(OR = exp(estimate),lower_CI = exp(conf.low),upper_CI = exp(conf.high)) %>%
  dplyr::select(model, term, OR, lower_CI, upper_CI, std.error, p.value)  %>%
  kable(caption = "Comparison of GLM and GLMM models") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))


# ----  Table: GLMM Odds Ratios ----
table_ready <- broom.mixed::tidy(rescue_glmer, conf.int = TRUE, effects = "fixed") %>%
  mutate(OR = exp(estimate), lower_CI = exp(conf.low), upper_CI = exp(conf.high)) %>%
  dplyr::select(term, OR, lower_CI, upper_CI, std.error, p.value) %>%
  mutate(term = case_when(
    term == "(Intercept)" ~ "Control (sgRNA_scramble)",
    term == "conditionsgRNA_thr" ~ "Thr Knockdown (sgRNA_thr)",
    term == "conditionsgThr_opsin2_20ug" ~ "Rescue (sgThr_opsin2_20ug)",
    TRUE ~ term))

format_pvalues_vec <- function(p_vec) {case_when(
    p_vec < 0.001 ~ format(p_vec, scientific = TRUE, digits = 2),
    TRUE ~ as.character(round(p_vec, 4))
  )
}

table_ready %>%
  mutate(p.value = format_pvalues_vec(p.value)) %>% 
  kbl(
    caption = "Table 1: Logistic Regression Analysis of Attachment Probability (GLMM)",
    digits = 3,
    booktabs = TRUE,
    col.names = c("Condition", "Odds Ratio", "95% CI Lower", "95% CI Upper", "Std. Error", "p-value"),
    align = "lccccc"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  column_spec(6, bold = (table_ready$p.value < 0.05)) #  bold  significant p-value 

```

|                        |                     |             |                                                                                |
|------------------------|---------------------|-------------|--------------------------------------------------------------------------------|
| **Contrast**           | **Odds Ratio (OR)** | **p-value** | **Biological Interpretation**                                                  |
| **KD vs. Control**     | **0.551**           | $0.0013$    | *Thr* knockdown reduces attachment odds by \~45%.                              |
| **Rescue vs. Control** | **3.986**           | $<0.0001$   | *Opsin2* increases attachment odds by nearly 4-fold over baseline.             |
| **Rescue vs. KD**      | **7.227**           | $<0.0001$   | **The Key Result:** Rescue has \>7x higher odds of success than the knockdown. |

```{r}
# ---- Forest Plot of Odds Ratios (effect size) ----
forest_plot <- ggplot(table_ready %>% filter(term != "Control (sgRNA_scramble)"), 
                      aes(x = OR, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  geom_errorbarh(aes(xmin = lower_CI, xmax = upper_CI), height = 0.2) +
  geom_point(size = 4, aes(color = term)) +
  scale_x_log10(breaks = c(0.5, 1, 3, 5)) +
  labs(
    title = "Effect Size (Odds Ratios)",
    x = "Odds Ratio (log scale)",
    y = ""
  ) +
  scale_color_manual(values = c("Thr Knockdown (sgRNA_thr)" = "#00AAAB", 
                                "Rescue (sgThr_opsin2_20ug)" = "#1970DB")) +
  custom_theme() +
  no_legend

forest_plot
```

## Binomial Confidence Intervals

Wilson CIs for asymmetry

```{r wilson_ci}

# ---- wilson CIs ----
opsn1_cas_rescue_BiCI.df <- opsn1_cas_rescue_condition.tidy %>%
  drop_na() %>%
  group_by(condition, status) %>%
  mutate(
    BiCI_low = binom.confint(count, total, methods = "wilson")$lower * 100,
    BiCI_high = binom.confint(count, total, methods = "wilson")$upper * 100
  )

# merge CI data with positioning info
CI_data <- opsn1_cas_rescue_summary.df %>%
  # align status names between datasets
  mutate(status = case_when(
    status == "attached_larvae" ~ "attached",
    status == "swimming_larvae" ~ "not_attached",
    TRUE ~ status
  )) %>%
  left_join(opsn1_cas_rescue_BiCI.df, by = c("condition", "status")) %>%
  # calculate error bar positions
  mutate(
    BiCI_low_position = y_start + BiCI_low - mean_percent,
    BiCI_high_position = y_start + BiCI_high - mean_percent
  )

CI_data %>%
  select(
    condition, status, mean_percent, y_start, BiCI_low, BiCI_high,
    BiCI_low_position, BiCI_high_position
  ) %>%
  kable(caption = "Wilson CI positions") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# eg repertive code but it works; to fix later 
# calculate pooled wilson CIs
wilson_stats <- opsn1_cas_rescue.df %>%
  drop_na() %>%
  mutate(total_larvae = swimming_larvae + attached_larvae) %>%
  group_by(condition) %>%
  summarise(
    attached_sum = sum(attached_larvae),
    total_sum = sum(total_larvae),
    .groups = "drop"
  ) %>%
  rowwise() %>%
  mutate(
    # calculate Wilson CI using the binom package
    res = list(binom.confint(attached_sum, total_sum, method = "wilson")),
    mean_prop = res$mean * 100,
    lower = res$lower * 100,
    upper = res$upper * 100
  ) %>%
  select(-res)

# simple bar plot with wilson CIs
ggplot() +
  # bar for the pooled proportion (attached only)
  geom_col(
    data = wilson_stats %>% filter(condition != "sgThr_opsin2_40ug"),
    aes(x = condition, y = mean_prop,fill = condition), width = 0.6, alpha = 0.7,
    
    color = "black" ) +
  scale_fill_manual(values = panel_a_colors) +
 
  # wilson confidence intervals (asymmetric)
  geom_errorbar(
    data = wilson_stats,
    aes(x = condition, ymin = lower, ymax = upper),
    width = 0.15, linewidth = 0.8, color = "black") +
  # individual replicate points (to show the "raw" variance)
  geom_jitter(
    data = filter(opsn1_cas_rescue.tidy, status == "attached_larvae" & condition != "sgThr_opsin2_40ug"),
    aes(x = condition, y = percent),
    shape = 21, fill = "white", color = "black", 
    size = 3, stroke = 1, width = 0.1) +
  # significance and N-labels
  geom_text(
    data = wilson_stats,
    aes(x = condition, y = upper + 5, label = paste0("n=", total_sum)),
    size = 3.5, vjust = 0
  ) +
  stat_pvalue_manual(stats_for_plot, y.position = 95, step.increase = 0.1, label = "label", tip.length = 0.02) +
  labs(y = "Larvae Attached (%)", x = NULL) +
  scale_y_continuous(limits = c(0, 110), breaks = seq(0, 100, 20)) +
  custom_theme() +
  no_legend


```

## Main Figure: Stacked Plot with Wilson CIs

```{r main_figure, fig.width=10, fig.height=8}
# ---- stacked plot + wilson CIs ----

rescue_plot_BiCI <- ggplot(
  opsn1_cas_rescue_summary.df,
  aes(x = condition, y = mean_percent, fill = status)
) +
  geom_col(width = 0.7, alpha = 0.8, color = "black") +  # stacked bars
  # error bars at attachment boundary (using calculated positions)
  geom_errorbar(
    data = CI_data %>% filter(status == "attached"),
    aes(ymin = BiCI_low_position, ymax = BiCI_high_position),
    width = 0.25,
    color = "black",
    linewidth = 0.8
  ) +
  # sample size labels
  geom_text(
    data = opsn1_cas_rescue_BiCI.df %>%
      filter(condition != "sgThr_opsin2_40ug") %>%
      group_by(condition) %>%
      summarize(total = first(total)),
    aes(x = condition, y = 110, label = paste0("n=", total)),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  scale_fill_manual(
    values = c("attached_larvae" = "#882222", "swimming_larvae" = "gray70"),
    labels = c("Attached", "Swimming")
  ) +
  labs(
    x = "",
    y = "Percentage of Larvae (%)",
    fill = "Attachment Status"
  ) +
  custom_theme() +
  theme(
    legend.position = "top",
    panel.grid.major.x = element_blank()
  ) +
  coord_cartesian(ylim = c(0, 115)) +
  # significance brackets and stars
  stat_pvalue_manual(
    data = data.frame(
      group1 = "sgRNA_scramble",
      group2 = c("sgRNA_thr", "sgThr_opsin2_20ug"),
      p.signif = c("**", "***"),  # GLMM p-values
      #== thi was my pug i previusly put *** and ***
      # but summary(rescue_glmer) is  
      # conditionsgRNA_thr          -0.5952     0.1848  -3.220  0.00128 ** 
    # conditionsgThr_opsin2_20ug   1.3827     0.1754   7.881 3.25e-15 ***
      status = c("attached", "attached"),
      y.position = c(110, 115)
    ),
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.6
  )

rescue_plot_BiCI
```

## Save Plots

```{r save_plots}
if (SAVE_OUTPUTS) {
  
  # save with cowplot - various sizes
  cowplot::save_plot(
    plot = rescue_plot_BiCI,
    base_height = 4,
    base_asp = 4 / 1.9,
    file = opath("photoreceptor_rescue_plot_a.pdf")
  )
  
  cowplot::save_plot(
    plot = rescue_plot_BiCI,
    base_height = 8,
    base_asp = 1.9,
    file = opath("photoreceptor_rescue_plot_b.pdf"),
    limitsize = FALSE
  )
  
  # save with CairoPDF
  width_in <- 40
  aspect_ratio <- 3/4
  height_in <- width_in * aspect_ratio
  
  CairoPDF(
    opath("photoreceptor_rescue_plot_cairo.pdf"),
    width = width_in,
    height = height_in
  )
  print(rescue_plot_BiCI)
  dev.off()
  
  # additional sizes
  cowplot::save_plot(
    plot = rescue_plot_BiCI,
    base_width = 2.4 * 5,
    base_height = 3.6 * 5,
    file = opath("photoreceptor_rescue_plot_final.pdf")
  )
  
  message("\nPlots saved to: ", out_dir)
}
```

## Session Info

```{r session_info}
sessionInfo()
```
