---
title: "Fig-S5H_thr_RT-qPCR_nkx2.1-KD"
author: "Andrea Mariossi"

format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    toc-depth: 5
    code-fold: false
    code-line-numbers: true
    page-layout: article
    grid:
      sidebar-width: 100px
      body-width: 2000px
      margin-width: 100px
      gutter-width: 3.5rem
execute:
  echo: true
  warning: false
  message: true
  
fig-width: 18
fig-height: 10
fig-align: center
fig-format: png
out-width: "90%"

editor_options:
  chunk_output_type: inline
---

## Setup

Inputs come either from my local folder or from the repository's folder (when downloaded from GitHub).

Outputs are written to the `output/` folder.

```{r setup}

rm(list = ls())

# package loading
suppressPackageStartupMessages({
  library(here)
  library(tidyverse)
  library(readxl)
  library(paletteer)
  library(rstatix)
  library(ggpubr)
  library(knitr)
  library(Cairo)
  library(lme4)
  library(lmerTest)
  library(emmeans)
  library(kableExtra)
  library(patchwork)
})
```

## Configuration

```{r config}

# ---- configuration ----
pick_pipeline <- "local"  # "local" or "github"

# ---- my paths (inputs+outputs) ----
file_path <- "/Users/andrea/Library/CloudStorage/GoogleDrive-andrea.mariossi@gmail.com/My Drive/4.paper_archive/2025_thyroid_pathway_&_metamorphosis/04.SciAdv_rev/source_data"

out_dir   <- file.path(file_path, "output")

# ---- GITHUB repo paths ----
repo_dir <- here()

# Fallback: check if repo exists in Downloads folder
if (!dir.exists(file.path(repo_dir, "data"))) {
  repo_dir_try <- path.expand("~/Downloads/thyroid_hormone_ciona-main")
  if (dir.exists(file.path(repo_dir_try, "data"))) {
    repo_dir <- repo_dir_try
  }
}

# ---- switch ----
if (pick_pipeline == "local") {
  data_dir <- file_path
  SAVE_OUTPUTS <- TRUE
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
  opath <- function(...) file.path(out_dir, ...)
  
} else if (pick_pipeline == "github") {
  data_dir <- file.path(repo_dir, "data")
  stopifnot(dir.exists(data_dir))
  
  SAVE_OUTPUTS <- TRUE
  out_dir <- file.path(repo_dir, "output")
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
  opath <- function(...) file.path(out_dir, ...)
  
} else {
  stop("pick_pipeline must be either 'local' or 'github'")
}

dpath <- function(...) file.path(data_dir, ...)

message("Pipeline: ", pick_pipeline)
message("Data dir (inputs): ", data_dir)
message("Outputs enabled: ", SAVE_OUTPUTS)
if (SAVE_OUTPUTS) message("Outputs dir: ", out_dir)

```

## Theme & Style

```{r theme}
# ---- theme & style ----
custom_theme <- function() {
  theme(
    axis.ticks = element_line(linewidth = 0.5),
    axis.ticks.length = unit(0.09, "cm"),
    legend.key.height = unit(20, "pt"),
    axis.line = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank(),
    strip.background = element_blank()
  )
}

# If using Arial font 
# library(showtext)
# library(sysfonts)
# sysfonts::font_add("Arial", regular = "/Library/Fonts/Arial.ttf")  # macOS 
# # sysfonts::font_add("Arial", regular = "C:/Windows/Fonts/arial.ttf")  # Windows
# showtext_auto()
# theme_set(theme_pubr(base_size = 13, base_family = "Arial") + custom_theme())

# ===== set theme =====
theme_set(theme_pubr(base_size = 13, base_family = "sans") + custom_theme())

# ===== extra bits =====
no_legend <- theme(legend.position = "none")
axis_text_45 <- theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

# ---- color palette ----
contPalette <- paletteer_c("ggthemes::Red-Green-Gold Diverging", 30)

# Stage ordering used throughout (edit if needed)
stage_order <- c("earTII", "midTII", "latTII", "latTIII", "larva")

# ---- pretty p-values for kable ----
format_pvalues <- function(df) {
  df %>%
    mutate(across(any_of(c("p.value", "p.value.x", "p.value.y", "adj.p.value", "p.adj")),
                  ~ case_when(
                    . < 0.001 ~ format(., scientific = TRUE, digits = 2),
                    TRUE ~ as.character(round(., 4))
                  )))
}
```

## Load Data

```{r load_data}
qpcr_file <- dpath("3_sgNkx2.1_thr_qPCR.xlsx")
stopifnot(file.exists(qpcr_file))

qPCR_GRN <- read_excel(qpcr_file)

qPCR_GRN <- qPCR_GRN %>%
  mutate(stage = factor(stage, levels = stage_order))

kable(head(qPCR_GRN), caption = "Raw qPCR Data (first 6 rows)")
```

## Quality Control: Raw Ct Values

```{r qc_raw_ct}
# calculating average Ct values for technical replicates
summarised_data <- qPCR_GRN %>%
  pivot_longer(cols = 5:10, names_to = "type", values_to = "Ct") %>%
  separate(type, into = c("gene", "tecRep"), sep = "_", convert = TRUE) %>%
  mutate(
    gene = as.factor(gene),
    bioRep = as.numeric(bioRep)
      #        bioRep = as.numeric(gsub("tecRep", "", bioRep))) |> 
  ) %>%
  group_by(gene, group, stage, hpf, bioRep) %>%
  summarise(mean_Ct = mean(Ct), .groups = "drop")

# plot average Ct values - actin & thr
ct_plot <- ggplot(summarised_data, aes(x = group, y = mean_Ct, colour = hpf)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_gradientn(colors = contPalette, name = "early to late hpf") +
  facet_wrap(~ gene, scales = "fixed") +
  labs(title = "Raw Ct Values: Actin and Thr") +
  custom_theme()

ct_plot

raw_ct_plot <- summarised_data %>%
  ggplot(aes(x = stage, y = mean_Ct, colour = hpf)) +
  geom_point(size = 2, alpha = 0.7, position = position_jitter(width = 0.12, height = 0)) +
  facet_wrap(~ gene, scales = "free_y") +
  scale_color_gradientn(colors = contPalette, name = "early → late hpf") +
  labs(
    x = "",
    y = "Mean Ct (technical replicates averaged)",
    title = "Raw Ct quality control"
  ) +
  custom_theme() +
  theme(legend.position = "right")

raw_ct_plot

```

### Confirm Actin Stability

Actin: Non-significant effects - reference gene is stable

```{r actin_stability}

#--------------------------------------------------------------------------------
# confirm actin Ct values don't change with treatment/development
#--------------------------------------------------------------------------------

actin_ct_plot_a <- ggplot(
  summarised_data %>% filter(gene == "actin"),
  aes(x = group, y = mean_Ct, colour = hpf)
) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_gradientn(colors = contPalette, name = "early to late hpf") +
  facet_wrap(~ gene, scales = "free_y") +
  labs(title = "Actin Ct Values (Reference Gene)") +
  custom_theme() 

actin_ct_plot_b <-  summarised_data %>% filter(gene == "actin") |> 
  ggplot(aes(x = stage, y = mean_Ct, colour = hpf)) +
  geom_point(size = 2, alpha = 0.7, position = position_jitter(width = 0.12, height = 0)) +
  facet_wrap(~ gene, scales = "free_y") +
  scale_color_gradientn(colors = contPalette, name = "early → late hpf") +
  labs(
    x = "",
    y = "Mean Ct (technical replicates averaged)",
    title = "Raw Ct quality control"
  ) +
  custom_theme() +
  theme(legend.position = "right")

actin_ct_plot_a + actin_ct_plot_b  + actin_ct_plot_a +
  facet_wrap(~ stage)  
  
# confirm actin Ct values don't change with treatment/development
anova_rawCt <- summarised_data %>%
  group_by(gene) %>%
  nest() %>%
  mutate(
    model = map(data, ~ aov(mean_Ct ~ group * stage, data = .x)),
    tidied = map(model, broom::tidy)
  ) %>%
  unnest(tidied) %>%
  select(gene, term, df, sumsq, meansq, statistic, p.value) %>%
  mutate(
    p.value = round(p.value, 6),
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )

anova_rawCt %>%
  format_pvalues() %>%
  kable(digits = 4, caption = "Quality Control: ANOVA raw Ct showing actin stability") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Interpretation:** Actin shows non-significant effects - reference gene is stable across conditions.

## ΔCt Calculation

```{r delta_ct}
# delta-Ct method: difference between Ct of housekeeping gene and target gene
goi_data <- summarised_data %>%
  filter(gene != "actin") %>%
  dplyr::rename("goiMean_Ct" = "mean_Ct")

ref_data <- summarised_data %>%
  filter(gene == "actin") %>%
  dplyr::rename("refMean_Ct" = "mean_Ct")

delta_Ct <- left_join(goi_data, ref_data, 
                      by = c("group", "hpf", "stage", "bioRep")) %>%
  dplyr::rename(
    gene_goi = gene.x,
    goi_Ct = goiMean_Ct,
    ref_Ct = refMean_Ct
  ) %>%
  relocate(group, bioRep, everything()) %>%
  mutate(
    delta_Ct = goi_Ct - ref_Ct,      # deltaCt
    mRNArel = 2^(-delta_Ct)           # relative mRNA
  )

# plot delta_Ct for each sample
delta_ct_plot <- ggplot(delta_Ct, aes(x = group, y = delta_Ct, colour = hpf)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_gradientn(colors = contPalette, name = "early to late hpf") +
  facet_wrap(~ gene_goi, scales = "fixed") +
  labs(
    x = "",
    y = "ΔCt (Target Gene - Reference)",
    title = "ΔCt Values Across Conditions"
  ) +
  custom_theme()

delta_ct_plot

# Quick visual sanity check of ΔCt
delta_ct_plot <- ggplot(delta_Ct, aes(x = group, y = delta_Ct, colour = hpf)) +
  geom_point(size = 2, alpha = 0.7, position = position_jitter(width = 0.1, height = 0)) +
  scale_color_gradientn(colors = contPalette, name = "early → late hpf") +
  facet_wrap(~ gene_goi, scales = "fixed") +
  labs(x = "", y = "ΔCt (Target - Reference)", title = "ΔCt values") +
  custom_theme()

delta_ct_plot

# look at distributions at each timepoint
delta_Ct %>%
  group_by(stage, hpf, group) %>%
  summarise(
    mean_mRNArel = mean(mRNArel),
    sd_mRNArel = sd(mRNArel),
    n = n(),
    .groups = "drop"
  ) %>%
  kable(digits = 6, caption = "mRNArel by Stage and Group") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Statistical Analysis

Does Nkx2.1 knockdown reduce Thr expression at each developmental stage?

\### Wilcoxon Tests

```{r wilcoxon_tests}

#------------------------------------
#  Wilcoxon tests
#------------------------------------

# wilcoxon tests for group comparisons at each stage
stat_wilcox <- delta_Ct %>%
  group_by(stage, hpf) %>%
  wilcox_test(mRNArel ~ group) %>%
  adjust_pvalue(method = "fdr") %>%
  add_significance() %>%
  mutate(
    test_type = "Mann-Whitney U (independent samples)",
    design = "Independent biological replicates across timepoints"
  )

# add effect sizes to show biological relevance
wilcox_with_effects <- delta_Ct %>%
  group_by(stage, hpf) %>%
  rstatix::wilcox_effsize(mRNArel ~ group) %>%
  left_join(stat_wilcox, by = c("stage", "hpf", "group1", "group2")) %>%
  select(stage, hpf, group1, group2, p.adj, p.adj.signif, effsize, magnitude)

# pretty table
wilcox_with_effects %>%
  kable(
    digits = 4,
    format.args = list(scientific = TRUE),
    caption = "Wilcoxon Test Results by Stage with Effect Sizes"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Fold reductions:**

-   7.7x reduction at earTII (0.0042 → 0.0005)
-   31.3x reduction at midTII (0.0075 → 0.0002)
-   76.9x reduction at latTII (0.0177 → 0.0002)
-   28.5x reduction at latTIII (0.0307 → 0.0011)
-   38.3x reduction at larva (0.0702 → 0.0018)

```{r}
# Summary stats for plotting
plot_summary <- delta_Ct %>%
  group_by(gene_goi, stage, hpf, group) %>%
  summarise(
    mean = mean(mRNArel, na.rm = TRUE),
    se   = sd(mRNArel, na.rm = TRUE) / sqrt(sum(!is.na(mRNArel))),
    .groups = "drop"
  )

# y-axis breaks on log scale
.yvals <- delta_Ct$mRNArel
.yvals <- .yvals[is.finite(.yvals) & .yvals > 0]
y_range <- range(.yvals, na.rm = TRUE)
y_breaks <- 10^(pretty(log10(y_range), n = 5))

ggplot() +
  geom_errorbar(
    data = plot_summary,
    aes(x = stage, ymin = pmax(mean - se, .Machine$double.eps), ymax = mean + se, group = group),
    position = position_dodge(width = 0.8),
    width = 0.2
  ) +
  geom_point(
    data = delta_Ct,
    aes(x = stage, y = mRNArel, colour = group),
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
    size = 2,
    alpha = 0.8
  ) +
  facet_wrap(~ gene_goi, scales = "free_y") +
  scale_y_log10(breaks = y_breaks, labels = scales::label_number()) +
  labs(
    x = "Stage\n(hpf)",
    y = "Relative expression (2^-ΔCt), log scale",
    title = "thr RT-qPCR after Nkx2.1 knockdown"
  ) +
  custom_theme() +
  theme(legend.position = "top") 

```

### ANOVA for Time Series Effect

If testing for the global pattern over the time series:

| gene_goi | term        | interpretation                             |
|----------|-------------|--------------------------------------------|
| thr      | group       | Does Nkx2.1 knockdown affect Thr?          |
| thr      | stage       | Does developmental stage affect Thr?       |
| thr      | group:stage | Does the knockdown effect depend on stage? |

```{r anova_timeseries}

#------------------------------------
# ANOVA for time series effect
#------------------------------------
# # if multiple genes tested asame time
# anova_results <- delta_Ct %>%
#   group_by(gene_goi) %>%
#   nest() %>%
#   mutate(
#     model = map(data, ~ aov(delta_Ct ~ group * stage, data = .x)),
#     tidied = map(model, broom::tidy)
#   ) %>%
#   unnest(tidied) 
# 
# anova_results %>%
#   select(-2, -3) %>%
#   mutate(p.value = sprintf("%.3e", p.value)) %>% 
#   kable(digits = 2, caption = "ANOVA Results for Time Series Effect")
# 
# anova_results

# two-way ANOVA for single gene (thr) expression
aov_result <- aov(mRNArel ~ group * stage,
                  data = delta_Ct %>% mutate(gene_goi = droplevels(gene_goi))
                  )

aov_result %>%
    broom::tidy() %>%
    format_pvalues() %>%
    kable(caption = "Two-way ANOVA: Group and Stage Effects") %>%
    kable_styling(bootstrap_options = c("striped", "hover"))

# tukey post-hoc tests
tukey_results <- TukeyHSD(aov_result)
#tukey_group <- TukeyHSD(aov_result, which = "group") # for just the group term
#tukey_interaction <- TukeyHSD(aov_result, which = "group:stage") # for jusy interaction term

tukey_results %>%
  broom::tidy() %>%
  mutate(p.adj = sprintf("%.3e", adj.p.value)) %>%
  filter(term == "group" | term == "stage") %>%
  kable(caption = "Tukey HSD Post-hoc Tests") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Extra QC: Trajectory Analysis

Does the knockdown alter the developmental trajectory (not just levels at each timepoint)?

\### Model Comparison: Interaction Effect

```{r model_comparison}

#------------------------------------
# First check interaction time and KD
#------------------------------------

# compare models with and without interaction
# null model: mRNArel ~ group + hpf (parallel lines)
null_model <- lm(mRNArel ~ group + hpf, data = delta_Ct)

# full model: mRNArel ~ group * hpf (different slopes)
full_model <- lm(mRNArel ~ group * hpf, data = delta_Ct)

model_comparison <- anova(null_model, full_model)

model_comparison %>%
  broom::tidy() %>%
  format_pvalues() %>%
  kable(caption = "Likelihood Ratio Test: Is Interaction Model Better?") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Interpretation:** Model comparison shows p = 4.205e-12 (interaction model is significantly better)

\### Slope Comparison

-   Are the linear slopes (rate of change over time) different between groups?
-   Does the knockdown change how fast expression increases/decreases over development?

```{r slope_comparison}

#------------------------------------
# test significant interaction time and KD loking at the slope
#------------------------------------

# simple linear model testing group × time interaction
trajectory_model <- lm(mRNArel ~ group * hpf, data = delta_Ct)
trajectory_anova <- anova(trajectory_model)

trajectory_anova %>%
  broom::tidy() %>%
  format_pvalues() %>%
  kable(caption = "Mixed Model ANOVA: Group × Time Interaction") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# compare slopes between groups
slope_comparison <- emtrends(trajectory_model, ~ group, var = "hpf")
slope_test <- pairs(slope_comparison)

slope_test %>%
  broom::tidy() %>%
  format_pvalues() %>%
  kable(caption = "Slope Comparison Between Groups") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Question:** Are the linear slopes (rate of change over time) different between groups? Does the knockdown change how fast expression increases/decreases over development?

\### Polynomial Model: Curvature

```{r polynomial_model}

#------------------------------------
# polynomial model to test curvature differences
#------------------------------------

poly_model <- lm(mRNArel ~ group * poly(hpf, 2), data = delta_Ct)
poly_anova <- anova(poly_model)

poly_anova %>%
  broom::tidy() %>%
  format_pvalues() %>%
  kable(caption = "Polynomial Model: Testing Curvature Differences") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Question:** Are the curves/shapes of the trajectories different? Does the knockdown change the curvature/non-linear pattern of development?

\### Trajectory Plot

```{r trajectory_plot, fig.width=12, fig.height=8}
trajectory_plot <- delta_Ct %>%
ggplot(aes(x = hpf, y = mRNArel, color = group, fill = group)) +
    geom_point(alpha = 0.7, size = 5, shape = 21, color = "black") +
    geom_smooth(
      method = "lm", formula = y ~ poly(x, 2),
      se = TRUE, alpha = 0.2, linewidth = 1.5
    ) +
    labs(
      subtitle = paste(
        "significant group x time interaction: p = 3.7e-11***\n",
        "Slopes significantly different: p < 0.0001***"
      ),
      x = "Hours Post Fertilization (hpf)",
      y = "Relative mRNA Expression (2^-ΔCt)",
      caption = paste(
        "Most probably other genes on top of Nkx2.1 regulate",
        "Thr expression late in development"
      )
    ) +
    scale_y_continuous(trans = "log10") +
    scale_color_manual(
      values = c("control" = "grey40", "sg Nkx2.1" = "#D55E00"),
      name = ""
    ) +
    scale_fill_manual(
      values = c("control" = "grey78", "sg Nkx2.1" = "#D55E00"),
      name = ""
    ) +
    custom_theme() +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
      plot.subtitle = element_text(hjust = 0.5, size = 12, face = "italic"),
      legend.position = "bottom"
    )

trajectory_plot
```

## Main Figure Panel

```{r main_figure, fig.width=12, fig.height=8}
# axis range & breaks
y_range <- range(delta_Ct$mRNArel, na.rm = TRUE)
y_breaks <- round(10^(seq(log10(y_range[1]), log10(y_range[2]), length.out = 5)), 4)
x_labels <- unique(delta_Ct$stage)
x_breaks <- unique(delta_Ct$hpf)
combined_labels <- paste(x_labels, "\n", x_breaks)

# determine the maximum y-value at each x break
max_y_values <- tapply(delta_Ct$mRNArel, delta_Ct$hpf, max)
spacing_constant <- 0.01
y_positions <- max_y_values + spacing_constant

# annotation df with p-values
annotations_df <- stat_wilcox %>%
  left_join(distinct(delta_Ct, stage, hpf), by = c("stage", "hpf")) %>%
  mutate(
    label = case_when(
      p.adj < 0.001 ~ "***",
      p.adj < 0.01 ~ "**",
      p.adj < 0.05 ~ "*",
      TRUE ~ "ns"
    ),
    y_pos = y_positions[as.character(hpf)]
  )

# mRNA expression plot with p-values annotations
mRNA_plot <- delta_Ct %>%
  ggplot(aes(x = hpf, y = mRNArel, group = group, color = group, fill = group)) +
  geom_point(alpha = 0.5, size = 6, shape = 21, color = "black") +
  # geom_smooth(aes(fill = group), size=2,method = "loess", se = TRUE) + # line + CI
  stat_summary(fun = mean, geom = "line", linewidth = 1.5) +
  stat_summary(fun.data = mean_sd, geom = "ribbon", alpha = 0.3, color = NA) +
  labs(x = "Developmental stage - hpf", 
       y = "Relative mRNA Expression Thr (2^-ΔCt)") +
  scale_y_continuous(trans = "log10", breaks = y_breaks) +
  scale_x_continuous(breaks = x_breaks, labels = combined_labels) +
  scale_color_manual(values = c("control" = "grey78", "sg Nkx2.1" = "#009E73"), name = "") +
  scale_fill_manual(values = c("control" = "grey78", "sg Nkx2.1" = "#009E73"), name = "") + 
  geom_text(data = annotations_df, 
            aes(x = hpf, y = y_pos, label = label), 
            size = 15, inherit.aes = FALSE, fontface = "bold") +
  theme(legend.position = c(0.2, 0.9),
          axis.title.x=element_blank()) +
  custom_theme() 
mRNA_plot
```

**Interpretation:**

-   Early stages: Knockdown suppresses thr developmental increase
-   Late stages: Some recovery/compensation appears

## Save Plots

```{r save_plots}
if (SAVE_OUTPUTS) {
  
  width_in <- 40
  aspect_ratio <- 3/4
  height_in <- width_in * aspect_ratio
  
  width_mm <- 4 * width_in / 25.4
  height_mm <- 4 * height_in / 25.4
  
  # save main figure with CairoPDF
  CairoPDF(
    opath("figS5H_thr_qPCR_nkx2.1-KD.pdf"),
    width = width_mm,
    height = height_mm
  )
  print(mRNA_plot)
  dev.off()
  
  # save with cowplot (aspect ratio)
  cowplot::save_plot(
    plot = mRNA_plot,
    base_height = height_in,
    base_asp = width_in / height_in,
    file = opath("figS5H_thr_qPCR_nkx2.1-KD_1.pdf")
  )
  
  cowplot::save_plot(
    plot = mRNA_plot,
    base_height = 0.9 * 3.71,
    base_asp = 0.9 * 1.618,
    file = opath("figS5H_thr_qPCR_nkx2.1-KD_2.pdf")
  )
  
  # save trajectory plot
  ggsave(
    opath("figS5H_trajectory_plot.pdf"),
    plot = trajectory_plot,
    width = 12, height = 8, device = cairo_pdf
  )
  
  message("\nPlots saved to: ", out_dir)
}
```

## Overall Conclusions

Nkx2.1 acts as a developmental modulator rather than simple on/off switch, regulating expression magnitude early while other factors contribute to late developmental timing. Nkx2.1 controls Thr expression dynamics during early-mid development, but late
expression is compensated by alternative regulatory mechanisms, explaining the partial rather than complete abolition observed in both qPCR and reporter assays.

1.  **Controls**\
    Reference gene (actin) validation confirms stable expression across all conditions (p \> 0.3), ensuring observed effects are biologically meaningful rather than technical artifacts.

<!-- -->

1.  **Highly Effective Knockdown**\
    Nkx2.1 CRISPR-Cas9 knockdown significantly reduces thyroid hormone receptor (thr) expression across all developmental stages (p \< 0.01 at each timepoint) with large effect sizes (Cohen's d = 0.83).

2.  **Stage-Dependent Effects**\
    thr expression shows strong developmental regulation in controls (p \< 0.0001), increasing \~17-fold from 9.3 to 17.5 hpf, but this pattern is disrupted by Nkx2.1 knockdown.

3.  The knockdown changes the developmental pattern, not just expression levels (GROUP × TIME interaction: p = 3.7e-11), with significantly different slopes and curvature between groups. This is consistent with the reporter assay showing **diminished but
    not abolished Thr** expression in **ΔNkx2.1 binding site** reporter data.

## Session Info

```{r session_info}
sessionInfo()
```
