---
title: "Fig-2D: T4 Window - Tail Regression"
author: "Andrea Mariossi"
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    toc-depth: 5
    code-fold: false
    code-line-numbers: true
    page-layout: article
    grid:
      sidebar-width: 100px
      body-width: 2000px
      margin-width: 100px
      gutter-width: 3.5rem
execute:
  echo: true
  warning: false
  message: true
  
fig-width: 18
fig-height: 10
fig-align: center
fig-format: png
out-width: "90%"

editor_options:
  chunk_output_type: inline
---

## Setup

-   **Inputs** come either from my local folder or from the repository’s folder (when downloaded from GitHub).
-   **Outputs** of GitHub are written to a new local folder named `output_rerun/`

```{r setup}
rm(list = ls())

# ---- packages ----
suppressPackageStartupMessages({
  library(here)
  library(tidyverse)
  library(readxl)
  library(paletteer)
  library(rstatix)
  library(ggpubr)
  library(knitr)
  library(Cairo)
  library(lme4)
  library(lmerTest)
  library(emmeans)
  library(kableExtra)
  library(patchwork)
  library(ggdist)
  library(gghalves)
  library(ggforce)
  library(forcats)
  library(cowplot)
  library(scales)
})
```

## Configuration

```{r config}
################################################################################
# ---- Configuration ----
################################################################################

pick_pipeline <- "my_local"  # "my_local" or "github"

# ---- My paths (inputs + outputs) ----
file_path <- "/Users/andrea/Library/CloudStorage/GoogleDrive-andrea.mariossi@gmail.com/My Drive/4.paper_archive/2025_thyroid_pathway_&_metamorphosis/04.SciAdv_rev/source_data"

# ---- Pipeline switch ----
if (pick_pipeline == "my_local") {
  
  # Inputs from my local folder
  data_dir <- file_path
  out_dir  <- file.path(file_path, "output")
  save_outputs <- TRUE
  
} else if (pick_pipeline == "github") {
  
  # ---- GitHub repo paths ----
  # By opening `thyroid_hormone_ciona.Rproj` in RStudio, `here()` finds the project root
  repo_dir <- here::here()
  
  # Fallback: if you downloaded the repo as a zip and unzipped it
  if (!dir.exists(file.path(repo_dir, "data"))) {
    repo_dir_try <- normalizePath(path.expand("~/Downloads/thyroid_hormone_ciona-main"), mustWork = FALSE)
    if (dir.exists(file.path(repo_dir_try, "data"))) {
      repo_dir <- repo_dir_try
    } else {
      stop("Cannot find repo 'data/' folder.\n",
           "  → Open the .Rproj file in RStudio, OR\n",
           "  → Set working directory to repo root with setwd(), OR\n",
           "  → Unzip thyroid_hormone_ciona-main.zip and update the path above.")
    }
  }
  
  data_dir <- file.path(repo_dir, "data")
  out_dir  <- file.path(repo_dir, "output_rerun")
  save_outputs <- TRUE
  
  # Uncomment line below to NOT save outputs when running from GitHub:
  # save_outputs <- FALSE
}

# ---- Path helper functions ----
dpath <- function(...) file.path(data_dir, ...)
opath <- function(...) file.path(out_dir, ...)

# Only create output dir when actually saving
if (save_outputs) dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# ---- Configuration summary ----
message("─────────────────────────────────────────")
message("Pipeline: ", pick_pipeline)
message("Data dir (inputs): ", data_dir)
if (exists("repo_dir")) message("Repo dir: ", repo_dir)
message("Save outputs: ", save_outputs)
if (save_outputs) message("Output dir: ", out_dir)
message("─────────────────────────────────────────")

```

## Theme & Style

```{r theme}
# ---- theme & style ----
custom_theme <- function() {
  theme(
    axis.ticks = element_line(linewidth = 0.5),
    axis.ticks.length = unit(0.09, "cm"),
    legend.key.height = unit(20, "pt"),
    axis.line = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank(),
    strip.background = element_blank()
  )
}

# If using Arial font 
# library(showtext)
# library(sysfonts)
# sysfonts::font_add("Arial", regular = "/Library/Fonts/Arial.ttf")  # macOS 
# # sysfonts::font_add("Arial", regular = "C:/Windows/Fonts/arial.ttf")  # Windows
# showtext_auto()
# theme_set(theme_pubr(base_size = 13, base_family = "Arial") + custom_theme())

# ===== set theme =====
theme_set(theme_pubr(base_size = 13, base_family = "sans") + custom_theme())

# ===== extra bits =====
no_legend <- theme(legend.position = "none")
axis_text_45 <- theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

# ---- color palette ----
contPalette <- paletteer_c("ggthemes::Red-Green-Gold Diverging", 30)

stage_order <- c("earTII", "midTII", "latTII", "latTIII", "larva")

# --- condition colors ---
condition_colors <- c(
  "control" = "#999999",
  "1"       = "#d9f0a3",
  "2"       = "#addd8e",
  "3"       = "#78c679",
  "5"       = "#41ab5d"
)

# --- colors for pie/donut charts ---
status_colors <- c("Attached" = "#ee3637", "Unattached" = "#747676")

# ---- pretty p-values for kable ----
format_pvalues <- function(df) {
  df %>%
    mutate(across(
      any_of(c("p.value", "p.value.x", "p.value.y", "adj.p.value")),
      ~ case_when(
        . < 0.001 ~ format(., scientific = TRUE, digits = 2),
        TRUE ~ as.character(round(., 4))
      )
    ))
}
```

## Load Data

```{r load_data}

T4_file <- dpath("8_T4_percentange_attachement.xlsx")
stopifnot(file.exists(T4_file))

T4_attach.df <- read_excel(T4_file)

# Clean and prepare data
data_tidy <- T4_attach.df %>%
  mutate(
    condition = factor(condition, levels = c("control", "1", "2", "3", "5")),
    fraction = attached / total,
    non_attached = total - attached
  ) %>%
  group_by(condition) %>%
  mutate(
    rep = row_number(),
    replicate = as.factor(rep),
    rep_offset = rep - mean(rep)
  ) %>%
  ungroup()

# Create long format for plotting
data_tidy_long <- data_tidy %>%
  pivot_longer(
    cols = c(attached, non_attached),
    names_to = "status",
    values_to = "count"
  )

# Calculate summary statistics
data_summary <- data_tidy %>%
  group_by(condition) %>%
  dplyr::summarise(
    mean_attached = mean(attached),
    mean_fraction = mean(fraction),
    median_fraction = median(fraction),
    sd_attached = sd(attached),
    sd_fraction = sd(fraction),
    total_count = sum(total),
    n_attached = sum(attached),
    n_unattached = total_count - n_attached,
    attached_pct = n_attached / total_count * 100,
    unattached_pct = n_unattached / total_count * 100,
    .groups = "drop"
  )

kable(data_summary, digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

## Descriptive Statistics

```{r descriptive_stats}

# ------------------------------------------------------------------------------
# ---- descriptive stats ----
# ------------------------------------------------------------------------------

# Total embryos overall
data_tidy %>%
  dplyr::summarise(total_embryos = sum(total, na.rm = TRUE)) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Number of replicates per condition
data_tidy %>%
  dplyr::summarise(n_replicates = n(), .by = condition) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Total embryos per condition
data_tidy %>%
  dplyr::summarise(total_embryos = sum(total, na.rm = TRUE), .by = condition) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Total attached/non-attached per condition
data_tidy %>%
  dplyr::summarise(
    attached_embryos = sum(attached, na.rm = TRUE),
    non_attached_embryos = sum(non_attached, na.rm = TRUE),
    .by = condition
  ) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# For adding n to plots
n_df <- data_tidy %>%
  dplyr::group_by(condition) %>%
  dplyr::summarise(total = sum(total, na.rm = TRUE), .groups = "drop") %>%
  dplyr::mutate(label = paste0("n=", total))

```

## Prepare Plot Data

```{r prepare_plot_data}

plot_data <- data_tidy %>%
  group_by(condition) %>%
  dplyr::summarise(
    total_attached   = sum(attached, na.rm = TRUE),
    total_unattached = sum(total - attached, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(total_attached, total_unattached),
    names_to = "status",
    values_to = "count"
  ) %>%
  mutate(status = dplyr::recode(
    status,
    total_attached = "Attached",
    total_unattached = "Unattached"
  )) %>%
  group_by(condition) %>%
  mutate(
    total   = sum(count),
    percent = 100 * count / total,
    fraction = percent / 100,
    ypos    = cumsum(percent) - 0.5 * percent,
    label   = sprintf("%s\n%.1f%%", status, percent)
  ) %>%
  ungroup()
```

## Main Figure - 2D - Pie Chart

```{r pie_chart, fig.width=12, fig.height=4}

# ------------------------------------------------------------------------------
# ---- Pie-chart data (percent WITHIN each condition) ----
# ------------------------------------------------------------------------------

# ---- Colors for pie segments ----
status_colors <- c("Attached" = "#ee3637", "Unattached" = "#747676")

# ---- Plot ----
pie_chart <- ggplot(plot_data, aes(x = "", y = percent, fill = status)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y", start = 0) +
  geom_text(aes(y = ypos, label = label), 
            color = "white", size = 4, fontface = "bold") +
  facet_wrap(~ condition, nrow = 1) +
  scale_fill_manual(values = status_colors) +
  labs(title = "Attachment Status by Condition", fill = "Status") +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11)
  ) +
  geom_text(
    data = plot_data %>% distinct(condition, total),
    aes(x = 1, y = 0, label = paste0("n=", total)),
    inherit.aes = FALSE,
    color = "black", size = 4, fontface = "bold"
  )

pie_chart
```

## Main Figure - 2D - Donut Chart

```{r donut_chart, fig.width=12, fig.height=4}

# ------------------------------------------------------------------------------
# Donut chart 
# ------------------------------------------------------------------------------
donut_chart <- ggplot(plot_data, aes(x = 2, y = percent, fill = status)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y", start = 0) +
  xlim(0.5, 2.5) +
  geom_text(aes(y = ypos, label = round(percent, 1)),
            color = "white", size = 4, fontface = "bold") +
  facet_wrap(~ condition, nrow = 1) +
  scale_fill_manual(values = status_colors) +
  labs(title = "Attachment Percentage by Condition", fill = "Status") +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11)
  ) +
  geom_text(
    data = plot_data %>% distinct(condition, total),
    aes(x = 0.55, y = 25, label = paste0("n=", total)),
    inherit.aes = FALSE,
    color = "black", size = 4, fontface = "bold"
  )

donut_chart
```

## QC Plots - Individual Replicates

```{r individual_replicates_plots}

# ---- stacked bar plot  ----
stacked_bar <- ggplot(plot_data, aes(condition, count, fill = status)) +
  geom_col(position = "fill", width = 0.7) +
  geom_text(
    aes(label = count),
    position = position_fill(vjust = 0.5),
    color = "white", fontface = "bold"
  ) +
  scale_fill_manual(values = status_colors) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportion of Attached vs Unattached by Condition",
    x = "Condition", y = "Percentage", fill = "Status"
  ) +
  custom_theme() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.major.x = element_blank()
  )

stacked_bar

# ---- dodged bar plot  ----
dodged_bar <- ggplot(plot_data, aes(condition, count, fill = status)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(
    aes(label = count),
    position = position_dodge(width = 0.7),
    vjust = -0.4, size = 3.5
  ) +
  scale_fill_manual(values = status_colors) +
  labs(
    title = "Count of Attached vs Unattached by Condition",
    x = "Condition", y = "Count", fill = "Status"
  ) +
  custom_theme() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

dodged_bar

# ---- heatmap per replicate ----
heatmap_plot <- ggplot(data_tidy_long, aes(condition, replicate, fill = fraction)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(
    aes(label = sprintf("%.1f%%", 100 * fraction)),
    color = "white", fontface = "bold"
  ) +
  scale_fill_gradient2(
    low = "#d7191c", mid = "#ffffbf", high = "#1a9641",
    midpoint = 0.6, labels = scales::percent_format()
  ) +
  labs(
    title = "Attachment Rate Heatmap by Condition and Replicate",
    x = "Condition", y = "Replicate", fill = "Attachment Rate"
  ) +
  custom_theme() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid = element_blank()
  ) +
  coord_equal()

heatmap_plot

# ---- points per replicate ----
points_by_rep <- data_tidy %>%
  filter(!is.na(fraction)) %>%
  ggplot(aes(x = condition, y = fraction)) +
  geom_errorbar(
    aes(ymin = fraction, ymax = fraction),
    width = 0.35, linewidth = 1.6, color = "black"
  ) +
  geom_point(
    aes(fill = condition),
    shape = 21, size = 3, color = "black", stroke = 0.8
  ) +
  geom_hline(
    yintercept = 0.5,
    linetype = "dashed", linewidth = 0.8, color = "grey70"
  ) +
  facet_wrap(~ replicate, ncol = 4) +
  scale_fill_manual(values = condition_colors) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Attachment Fraction by Condition for Each Replicate",
    x = "Condition", y = "Attached Fraction"
  ) +
  custom_theme() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

points_by_rep

```

### Combined QC Figure

```{r combined_qc, fig.width=18, fig.height=16}
#| fig-height: 16
#| fig-width: 18

# ---- multi-panels figure  ----
combined_plot <-
  ((heatmap_plot + labs(title = NULL)) | 
   (points_by_rep + labs(title = NULL))) /
  ((stacked_bar + labs(title = NULL)) | 
   (dodged_bar + labs(title = NULL))) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a") &
  theme(plot.title = element_blank())

combined_plot

# ((heatmap_plot + labs(title = NULL)) | (points_by_rep + labs(title = NULL))) /
# (stacked_bar + labs(title = NULL)) /
# (dodged_bar + labs(title = NULL)) +
# plot_layout(guides = "collect") +
# plot_annotation(tag_levels = "a") &
# theme(plot.title = element_blank())

```

## QC Plots - aggregate

### QC Plots - Violin Plot

```{r violin_plot}
# --- Replicate shape ---
n_reps <- nlevels(data_tidy$replicate)

# Create shape values based on number of replicates
if (n_reps <= 5) {
  shape_values <- 21:(20 + n_reps)
} else if (n_reps <= 7) {
  shape_values <- c(21:25, 3, 4)
} else {
  shape_values <- 1:n_reps
}

# --- fast violin plot ---
violin_plot <- ggplot(data_tidy, aes(condition, fraction, fill = condition)) +
  geom_violin(alpha = 0.6, width = 0.8, trim = TRUE) +
  geom_boxplot(width = 0.2, alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.1, height = 0, size = 2.5, alpha = 0.7) +
  stat_summary(
    fun = mean, geom = "point",
    shape = 23, size = 4, fill = "white", color = "black"
  ) +
  stat_summary(
    fun = mean, geom = "text",
    aes(label = paste0(round(after_stat(y) * 100, 1), "%")),
    vjust = -2, size = 3.5, fontface = "bold"
  ) +
  scale_fill_manual(values = condition_colors) +
  scale_y_continuous(limits = c(0.35, 1), labels = scales::percent_format()) +
  labs(x = "Condition", y = "Attached Fraction") +
  custom_theme() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.minor = element_blank()
  )

violin_plot
```

## 

### QC Plots - Half Violin with Sina Points

```{r main_figure, fig.width=10, fig.height=8}

# --- half & points plot ---
main_p <- ggplot(data_tidy, aes(condition, fraction)) +
  # half violin
  gghalves::geom_half_violin(
    aes(fill = condition),
    side = "l", adjust = 0.7, alpha = 0.7, width = 1.1,
    position = position_nudge(x = -0.15),
    show.legend = FALSE
  ) +
  # half boxplot
  gghalves::geom_half_boxplot(
    side = "l", width = 0.2,
    fill = "white", color = "black",
    outlier.shape = NA, errorbar.draw = FALSE,
    position = position_nudge(x = -0.15)
  ) +
  # mean/interval pointinterval adjusted to center
  ggdist::stat_pointinterval(
    position = position_nudge(x = -0.15),
    show.legend = FALSE
  ) +
  # Sina points with replicate shapes
  ggforce::geom_sina(
    aes(fill = condition, shape = replicate),
    color = "black", alpha = 0.8, size = 5,
    bin_limit = 1, maxwidth = 0.2,
    show.legend = c(fill = FALSE, shape = TRUE)
  ) +
  # Sample size labels
  geom_text(
    data = n_df,
    aes(x = condition, y = 0.98, label = label),
    inherit.aes = FALSE,
    size = 4, color = "black"
  ) +
  # Scales & theme
  scale_fill_manual(values = condition_colors, guide = "none") +
  scale_shape_manual(values = shape_values[seq_len(n_reps)], name = "Rep") +
  scale_y_continuous(
    limits = c(0.35, 1),
    labels = scales::percent_format(),
    oob = scales::squish
  ) +
  labs(x = "Condition", y = "Attached Fraction") +
  custom_theme() +
  theme(legend.position = c(0.2, 0.8))

main_p
```

#### extra layout

```{r}
# --- figure plot ---
p <- ggplot(data_tidy, aes(condition, fraction)) +
  # Half violin
  gghalves::geom_half_violin(
    aes(fill = condition),
    side = "l", adjust = 0.7, alpha = 0.7, width = 1.1,
    position = position_nudge(x = -0.15),
    show.legend = FALSE
  ) +
  # Half boxplot
  gghalves::geom_half_boxplot(
    side = "l", width = 0.2,
    fill = "white", color = "black",
    outlier.shape = NA, errorbar.draw = FALSE,
    position = position_nudge(x = -0.15)
  ) +
  # mean/interval pointinterval adjusted to center
  ggdist::stat_pointinterval(
    position = position_nudge(x = -0.15),
    show.legend = FALSE
  ) +
  # scales & labels
  scale_fill_manual(values = condition_colors, guide = "none") +
  scale_y_continuous(limits = c(0.35, 1), labels = scales::percent_format(), oob = scales::squish) +
  labs(x = "Condition", y = "Attached Fraction") +
  custom_theme()

p + geom_point(aes(fill = condition),shape = 21,color = "black",size = 3,    
    # position = position_dodge(width = 0.2)
    )  +

  # add n labels near the top
  geom_text(
    data = n_df,
    aes(x = condition, y = 0.98, label = label),
    inherit.aes = FALSE,
    size = 4,
    color = "black"
  )
  
p +  
  # Sina points pushed right
    ggforce::geom_sina(
    aes(fill = condition),
    shape = 21,
    # aes(fill = condition, shape = replicate),
    color = "black",alpha = 2,
    size = 5, bin_limit = 1, maxwidth = 0.2,
    show.legend = c(fill = FALSE, shape = TRUE) 
  ) +
  scale_shape_manual(values = shape_values[seq_len(n_reps)], name = "Rep") +
  theme(legend.position = c(0.2,0.8)) +

  # add n labels near the top
  geom_text(
    data = n_df,
    aes(x = condition, y = 0.98, label = label),
    inherit.aes = FALSE,
    size = 4,
    color = "black"
  )

```

## Save Plots

```{r save_plots}
if (save_outputs) {
  
  # Save pie chart
  ggsave(
    opath("fig2c_pie_chart.pdf"),
    plot = pie_chart,
    width = 12, height = 4, device = cairo_pdf
  )
  
  # Save donut chart
  ggsave(
    opath("fig2c_donut_chart.pdf"),
    plot = donut_chart,
    width = 12, height = 4, device = cairo_pdf
  )
  
  # Save QC combined plot
  ggsave(
    opath("fig2c_qc_combined.pdf"),
    plot = combined_plot,
    width = 18, height = 16, device = cairo_pdf
  )
  
  # Save violin plot
  ggsave(
    opath("fig2c_violin_plot.pdf"),
    plot = violin_plot,
    width = 8, height = 6, device = cairo_pdf
  )
  
  # Save main figure
  ggsave(
    opath("fig2c_main_plot.pdf"),
    plot = main_p,
    width = 10, height = 8, device = cairo_pdf
  )
  
  message("\nPlots saved to: ", out_dir)
}
```

## Session Info

```{r session_info}
sessionInfo()
```
