---
title: "Thyroid Inhibitor Effects on Metamorphosis"
author: "Andrea Mariossi"
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    toc-depth: 5
    code-fold: false
    code-line-numbers: true
    page-layout: article
    grid:
      sidebar-width: 100px
      body-width: 2000px
      margin-width: 100px
      gutter-width: 3.5rem
execute:
  echo: true
  warning: false
  message: true
fig-width: 12
fig-height: 8
fig-align: center
fig-format: png
editor_options:
  chunk_output_type: inline
---

## Experimental Design

**Objective:** Test the effect of thyroid hormone inhibitors (goidrogens) on metamorphosis by measuring tail regression.

**Design:**

\- **Conditions:** 4 (Control, Perchlorate, Methimazole, Iopanoic Acid)

\- **Biological replicates:** Multiple independent experiments (biorep)

\- **Response variable:** regression percentages were grouped into four bins (0–30%, 30–50%, 50–80%, and 80–100%)

\- **Measurement:** Count of larvae in each tail regression category per condition × bioreps.

## Setup

```{r setup}
rm(list = ls())

# ---- packages ----
suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
  library(ggpubr)
  library(knitr)
  library(kableExtra)
  library(rstatix)
  library(emmeans)
  library(Cairo)
  library(patchwork)
  library(scales)
})
```

## Data path

```{r data-loadin}
# ---- paths ----
data_dir <- file.path("..", "source_data")
out_dir  <- file.path("..", "output")

# Uncomment for local absolute path:
# data_dir <- "/Users/andrea/Library/CloudStorage/GoogleDrive-andrea.mariossi@gmail.com/My Drive/4.paper_archive/2025_thyroid_pathway_&_metamorphosis/04.SciAdv_rev/source_data/"

SAVE_OUTPUTS <- FALSE
if (!dir.exists(out_dir)) dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# helpers
dpath <- function(...) file.path(data_dir, ...)
opath <- function(...) file.path(out_dir, ...)

```

## Theme & Style

```{r theme-setup}
# ---- theme & style ----
custom_theme <- function() {
  theme(
    axis.ticks = element_line(linewidth = 0.5),
    axis.ticks.length = unit(0.09, "cm"),
    legend.key.height = unit(20, "pt"),
    axis.line = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank(),
    strip.background = element_blank()
  )
}

theme_set(theme_pubr(base_size = 13, base_family = "Helvetica") + custom_theme())

# ---- condition order ----
cond_levels <- c("Control","Cond_1","Cond_2","Cond_3","Cond_4","Cond_5")

# ---- color palette ----

# Custom color palettes
palette_teal_orange <- scale_fill_manual(values = c(
  "#66c2a5",  
  "#fc8d62", 
  "#8da0cb", 
  "#DDCC77"  
))

# For tail regression categories
tail_colors <- c(
  "0-30%"    = "#f37749", 
  "30-50%"   = "#faf7b4",
  "50-80%"   = "#70d2ca",
  "80-100%"  = "#f3f3f4"  
)

# For conditions
condition_colors <- c(
  "control"        = "#e78ac3",
  "perchlorate"    = "#117733",
  "methimazole"    = "#332288",
  "iopanoic acid"  = "#88CCEE"
)

# ---- p-value formatting function ----
format_pvalues <- function(df) {
  df %>%
    mutate(across(
      any_of(c("p.value", "p.adj", "adj.p.value")),
      ~ case_when(
        . < 0.001 ~ format(., scientific = TRUE, digits = 2),
        TRUE ~ as.character(round(., 4))
      )
    ))
}
```

## Data load & Preprocessing

```{r data-tidy}}
# ---- dataset ----
data_file <- "10_inhibitors_metamorphosis_day1.xlsx"
stopifnot(file.exists(dpath(data_file)))

# ---- tidy df ----
drugInhibitor.df <- read_excel(dpath(data_file)) %>%
  rename_with(tolower) %>%
  mutate(
    condition = factor(condition, levels = c("control", "perchlorate", "methimazole", "iopanoic acid")),
    tail.regression = factor(tail.regression, levels = c("0-30%", "30-50%", "50-80%", "80-100%")),
    biorep = factor(biorep)
  )

cat("\n=== DATA STRUCTURE ===\n")
cat("Conditions:", levels(drugInhibitor.df$condition), "\n")
cat("Tail regression categories:", levels(drugInhibitor.df$tail.regression), "\n")
cat("Number of biological replicates:", n_distinct(drugInhibitor.df$biorep), "\n")
cat("Total observations:", nrow(drugInhibitor.df), "\n\n")

# Ppeview
drugInhibitor.df %>%
  head(10) %>%
  kable(caption = "Drug inhibitor data (first 10 rows)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# ---- n embryos x condition  ----
drugInhibitor.df |>
  group_by(condition) |>
  summarise(total_embryos = sum(n, na.rm = TRUE), .groups = "drop")  %>%
  kable(caption = "n embryos x condition") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))


```

## Calculate Percentages

```{r data-transformation}
# ---- percentages within each condition × biorep ----
drug_percent <- drugInhibitor.df %>%
  group_by(condition, biorep) %>%
  mutate(total = sum(n),
          n_reps = n(),
          se_n =sd(n) / sqrt(n_reps), 
         percent = n / total * 100) %>%
  ungroup()

drug_percent %>%
  select(condition, biorep, tail.regression, n, total, percent) %>%
  kable(caption = "Percentages by condition, biorep, and tail regression category") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# ---- summary counts ----
counts_stats <- drug_percent %>%
  group_by(condition, tail.regression) %>%
  summarise(mean_n = mean(n),
            range_n = paste0(min(n), "–", max(n)),
            se_n   = sd(n) / sqrt(n()),
            .groups = "drop")
counts_stats

# ---- summary statistics ----
summary_stats <- drug_percent %>%
  group_by(condition, tail.regression) %>%
    summarise(
    n_reps = n(), # number of observations
    mean_pct = mean(percent), # Mean percentage
    sd_pct = sd(percent), # Standard deviation
    se_pct = sd_pct / sqrt(n_reps), #  Standard error
    min_pct = min(percent),
    max_pct = max(percent),
    .groups = "drop"
  )

summary_stats %>%
  kable(caption = "Summary statistics: mean % ± SE per condition and tail regression category",
        digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

## Statistical Analysis

### Chi-Square Analysis: Overall Distribution Test

**Question:** Do the drugs change the overall tail-regression distribution across bins (0–30, 30–50, 50–80, 80–100)?

```{r chisq-overall}

# ---- chi-square test: condition × tail regression ----
# contingency table (aggregate across bioreps)
count_table <- drugInhibitor.df %>%
  group_by(condition, tail.regression) %>%
  summarise(total_n = sum(n), .groups = "drop") %>%
  pivot_wider(names_from = tail.regression, values_from = total_n, values_fill = 0)

count_table %>%
  kable(caption = "Contingency table: condition × tail regression") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Convert to matrix for chi-square test
count_matrix <- count_table %>%
  select(-condition) %>%
  as.matrix()
rownames(count_matrix) <- count_table$condition
 
# Overall chi-square test
chisq_result <- chisq.test(count_matrix)
cat("\n=== Chi-Square Test ===\n")
print(chisq_result)

# Effect size (Cramér's V)
V_overall <- sqrt(as.numeric(chisq_result$statistic) /
                  (sum(count_matrix) * min(nrow(count_matrix) - 1, ncol(count_matrix) - 1)))
cat("\nCramér's V:", round(V_overall, 3), "\n")

# Standardized residuals (which cells drive the effect)
stdres_overall <- as.data.frame(as.table(chisq_result$stdres)) %>%
  rename(condition = Var1, bin = Var2, stdres = Freq) %>%
  arrange(desc(abs(stdres)))

stdres_overall %>%
  kable(caption = "Standardized residuals (sorted by magnitude)", digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Residual Heatmap
ggplot(stdres_overall, aes(x = bin, y = condition, fill = stdres)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(stdres, 1)), size = 5) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(
    x = "Tail Regression Bin",
    y = "Condition",
    fill = "Std. Residual",
    title = "Standardized Residuals: Which cells drive the overall chi-square?"
  ) 

```

## Bin Proportions

```{r}
# Calculate top-bin proportion and ordinal mean using bin midpoints
bin_midpoints <- c("0-30%" = 15, "30-50%" = 40, "50-80%" = 65, "80-100%" = 90)

descriptive_summary <- as.data.frame(count_matrix) %>%
  tibble::rownames_to_column("condition") %>%
  mutate(
    n = rowSums(across(where(is.numeric))),
    top_bin_prop = `80-100%` / n,
    mean_pct = (
      `0-30%` * bin_midpoints["0-30%"] +
      `30-50%` * bin_midpoints["30-50%"] +
      `50-80%` * bin_midpoints["50-80%"] +
      `80-100%` * bin_midpoints["80-100%"]
    ) / n
  ) %>%
  select(condition, n, mean_pct, top_bin_prop) %>%
  arrange(desc(mean_pct))

descriptive_summary %>%
  mutate(
    mean_pct = round(mean_pct, 1),
    top_bin_prop = scales::percent(top_bin_prop, accuracy = 0.1)
  ) %>%
  kable(caption = "Descriptive summary: ordinal mean & proportion with 80-100% regression") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

| Condition     | Mean %  | 80-100% bin | Interpretation                           |
|---------------|---------|-------------|------------------------------------------|
| Control       | \~78.6% | 69.8%       | Normal metamorphosis                     |
| Perchlorate   | \~42.9% | 2.8%        | **Severe inhibition** (strongest effect) |
| Methimazole   | \~59.0% | 9.2%        | Moderate inhibition                      |
| Iopanoic Acid | \~61.1% | 11.6%       | Moderate inhibition                      |

### Pairwise Chi-Square Tests

**Question:** Which pairs of conditions differ in their bin distributions?

```{r pairwise-chisq}

# -----------------------
# Pairwise overall-shift tests (each is 2x4)

pairs <- combn(rownames(count_matrix), 2, simplify = FALSE)

pairwise_results <- map_dfr(pairs, function(p) {
  sub <- count_matrix[p, , drop = FALSE]
  tst <- chisq_test(sub)
  
  tibble(
    group1 = p[1],
    group2 = p[2],
    statistic = tst$statistic,
    df = tst$df,
    p = tst$p,
    cramer_v = cramer_v(sub),
    method    = tst$method
  )
}) %>%
  mutate(p.adj = p.adjust(p, method = "holm")) %>%
  arrange(p.adj)

## this if you dont use rstatix package
  # tst <- chisq.test(sub)               # use fisher.test(sub) if you prefer. to kee pfor preference
  # tibble(
  #   group1 = p[1],
  #   group2 = p[2],
  #   statistic = unname(tst$statistic),
  #   p = tst$p.value,
  #   method = tst$method
  # )
  
pairwise_results %>%
  format_pvalues() %>%
  kable(caption = "Pairwise chi-square tests (Holm-adjusted)", digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

## QC

### Dispersion in Biological Replicate

```{r}
#| label: qc-biorep
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 14

# Counts by condition
p_rep_a <- ggplot(drugInhibitor.df, aes(x = tail.regression, y = n, fill = condition)) +
  geom_col(position = "dodge") +
  facet_wrap(~ biorep) +   # split by replicate
  labs(x = "", y = "Count", fill = "Condition") +
  scale_fill_manual(values = condition_colors, name = "Condition") +
  theme(legend.position = "bottom")

  # pd <- position_dodge(width = 0.9)

# ggplot(drug_percent, aes(x = interaction(tail.regression, biorep, sep = " | "),
#                          y = n, fill = condition)) +
#   geom_col(position = pd, width = 0.9) +
#   labs(x = "", y = "Count", fill = "Condition") +
#   scale_fill_manual(values = condition_colors, name = "Condition") +
#   theme(legend.position = "bottom") +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))

# % by condition
p_rep_b <- ggplot(drug_percent, aes(x = condition, y = percent, fill = tail.regression)) +
  geom_col(position = "stack", color = "black") +
  facet_wrap(~ biorep, ncol =2) +
  scale_fill_manual(values = tail_colors, name = "Tail Regression") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(y = "% of Larvae") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),legend.position = "bottom")

p_rep_c <- drug_percent %>%
  ggboxplot(x = "condition", 
            y = "percent",
            fill = "tail.regression",
            palette = "RdYlBu",
            ylab = "% Tail Regression",
            xlab = "") +
  scale_y_continuous(limits = c(0, 100)) +
  scale_fill_manual(values = tail_colors, name = "Tail Regression") +
  theme(legend.position = "none")

### Confidence Intervals for Proportions
# Calculate Wilson 95% confidence intervals
df_ci <- as.data.frame(as.table(count_matrix)) %>%
  rename(condition = Var1, bin = Var2, n = Freq) %>%
  group_by(condition) %>%
  mutate(
    N = sum(n),
    prop = n / N,
    z = qnorm(0.975),
    denom = 1 + (z^2) / N,
    center = (prop + (z^2) / (2 * N)) / denom,
    half = (z * sqrt((prop * (1 - prop)) / N + (z^2) / (4 * N^2))) / denom,
    lo = pmax(0, center - half),
    hi = pmin(1, center + half)
  ) %>%
  ungroup()

p_ci <- ggplot(df_ci, aes(x = condition, y = prop, fill = bin)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = lo, ymax = hi),
    position = position_dodge(width = 0.9),
    width = 0.2
  ) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = tail_colors) +
  labs(x = "", y = "Proportion (±95% CI)", fill = "Tail Regression",
    # title = "Bin Proportions with 95% Confidence Intervals"
  ) +
  theme(legend.position = "none")

p_rep_a + p_rep_b + p_rep_c + p_ci +
  plot_annotation(title = "Counts & %: tail Regression distribution by Biological Replicate",
                   theme = theme(plot.title = element_text(size = 22, hjust = 0.5, face="bold"))) +
   plot_layout(guides = 'collect')
```

### Overall effect - Proportional bar plot

```{r}
#| label: stacked-proportions
#| fig-width: 10
#| fig-height: 6

# Counts by condition
pd <- position_dodge(width = 0.9)
p_cond_a <- ggplot(counts_stats, aes(x = tail.regression, y = mean_n, fill = condition)) +
  geom_col(position = pd, width = 0.9) +
  geom_errorbar(aes(ymin = pmax(0, mean_n - se_n), ymax = mean_n + se_n, group = condition),position = pd, width = 0.2) +
  labs(x = "", y = "Count", fill = "Condition") +
  scale_fill_manual(values = condition_colors, name = "Condition") +
  theme(legend.position = "top")

#-------------------------------------------------------------------------------
# Summary Plot with Error Bars x counts
p_cond_b <- drugInhibitor.df %>% 
  ggbarplot(x = "condition", 
            y = "n", 
            add = "mean_se", 
            color = "tail.regression", 
            palette = "RdYlBu",
            label = TRUE, 
            lab.col = "black", 
            lab.pos = "in")

p_cond_a + p_cond_b
# # scale to 100% using ggbarplot
# drugInhibitor.df %>%
#   group_by(condition, biorep) %>%
#   mutate(percent = round(n * 100 / sum(n, na.rm = TRUE), 1)) %>%
#   ggbarplot(x = "condition", y = "percent", add = "mean_se", error.plot = "errorbar", width = 0.5,
#             ylab = "% tail regression", xlab = "Drug", color = "black", fill = "tail.regression", 
#             # color = "tail.regression", 
#             palette = "RdYlBu", label = TRUE, lab.col = "black", lab.size = 5, lab.pos = "in", 
#             lab.nb.digit = 1) +
#   scale_y_continuous(labels = scales::percent_format(scale = 1), breaks = c(25, 50, 75, 100)) +
#   theme(legend.position = "bottom") 
```

## Figure S10 A: Stacked Bar Plot with Error Bars

```{r main-stacked-plot}

# cumulative positions for error bars
summary_for_plot <- summary_stats %>%
  group_by(condition) %>%
  arrange(tail.regression) %>%
  mutate(
    # Cumulative positions (reversed for proper stacking)
    ymax_cumul = cumsum(mean_pct),
    ymin_cumul = lag(ymax_cumul, default = 0),
    ymid_cumul = (ymin_cumul + ymax_cumul) / 2
  ) %>%
  ungroup()

#  p-values for each treatment vs control from pairwise_results
pval_annotations <- pairwise_results %>%
  filter(group1 == "control") %>%
  mutate(
    label = case_when(
      p.adj < 0.001 ~ "***",
      p.adj < 0.01 ~ "**",
      p.adj < 0.05 ~ "*",
      TRUE ~ "ns"
    ),
    condition = group2,
    y_pos = 105
  ) %>%
  select(condition, label, y_pos)


# stacked bar plot
p_main <- ggplot(summary_for_plot, aes(x = condition, fill = tail.regression)) +
  # Stacked bars
  geom_col(
    aes(y = mean_pct),
    position = position_stack(reverse = TRUE),
    color = "black",
    width = 0.6,
    linewidth = 0.3
  ) +
  # Error bars
  geom_errorbar(
    aes(
      ymin = ymin_cumul + mean_pct - se_pct,
      ymax = ymin_cumul + mean_pct + se_pct
    ),
    width = 0.18,
    color = "black",
    linewidth = 0.5
  ) +
  # Labels
  geom_text(
    aes(y = ymid_cumul, label = sprintf("%.1f%%", mean_pct)),
    color = "black",
    size = 4
  ) +
   # Add significance asterisks
  geom_text(
    data = pval_annotations,
    aes(x = condition, y = y_pos, label = label, fill = NULL),
    size = 6,
    fontface = "bold"
  ) +
  # Styling
  scale_fill_manual(values = tail_colors, name = "Tail regression progression") +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    # labels = scales::percent_format(scale = 1),
    breaks = seq(0, 100, 25),
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_x_discrete(
    breaks = c("control", "perchlorate", "methimazole", "iopanoic acid"),
    labels = c("Control", "Perc", "MMI", "IPA")
  ) +
  labs(y = "Proportion") +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(size = 12, margin = margin(t = 2)),
    panel.spacing.x = unit(0.1, "cm"),  # Tighter than 0.2
    aspect.ratio = 1.2  # Original: 1. Makes plot wider/shorter
  )
  # coord_cartesian(clip = "off") +
  # custom_theme() 

p_main 
 
```

#### Extra plot on High Regression (80-100%)

```{r  anova-plot-high-regression}
#| label: boxplot-high-regression
#| fig-width: 10
#| fig-height: 6


high_regression <- drug_percent %>%
  filter(tail.regression == "80-100%")

# ANOVA
fit_high <- lm(percent ~ condition, data = high_regression)

cat("\n=== ANOVA: % with 80-100% Tail Regression ===\n")
print(anova(fit_high))

# Post-hoc contrasts vs control
emm_high <- emmeans(fit_high, ~ condition)
contrasts_high <- summary(
  contrast(emm_high, method = "trt.vs.ctrl", ref = "control"),
  adjust = "BH",
  infer = TRUE
)

contrasts_high %>%
  as.data.frame() %>%
  format_pvalues() %>%
  kable(caption = "Contrasts vs Control (80-100% regression; BH adjusted)", digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Comparison groups for statistics
comparisons <- list(
  c("control", "perchlorate"),
  c("control", "methimazole"),
  c("control", "iopanoic acid")
)

# High regression bin (80-100%)
ggplot(high_regression, aes(x = condition, y = percent, fill = condition)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.6) +
  scale_fill_manual(values = condition_colors) +
  labs(
    x = "Treatment Condition",
    y = "% Larvae with 80-100% Tail Regression",  ) +
  stat_compare_means(
    comparisons = comparisons,
    method = "t.test",
    label = "p.signif"
  ) +
  guides(fill = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Save Outputs

```{r save-outputs}
if (SAVE_OUTPUTS) {
  # ---- figures ----
  ggsave(opath("inhibitor_stacked_bars.svg"), p_stacked, width = 10, height = 6)
  ggsave(opath("inhibitor_faceted_bioreps.svg"), p_faceted, width = 14, height = 8)
  ggsave(opath("inhibitor_high_regression_boxplot.svg"), p_box_high, width = 10, height = 6)
  ggsave(opath("inhibitor_combined_figure.svg"), combined_plot, width = 14, height = 10)
  
  # PDFs
  CairoPDF(opath("inhibitor_combined_figure.pdf"), width = 14, height = 10)
  print(combined_plot)
  dev.off()
  
  message("\nOutputs saved to: ", out_dir)
}
```

## Session Info

```{r}
sessionInfo()
```
