---
title: "Fig-1I_Nkx2.1 Mutant Thr Reporter"
author: "Andrea Mariossi"
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    toc-depth: 5
    code-fold: false
    code-line-numbers: true
    page-layout: article
    grid:
      sidebar-width: 100px
      body-width: 2000px
      margin-width: 100px
      gutter-width: 3.5rem
execute:
  echo: true
  warning: false
  message: true
fig-width: 18
fig-height: 10
fig-align: center
fig-format: png
out-width: "90%"
editor_options:
  chunk_output_type: inline
---

-   **Inputs** come either from my local folder or from the repository’s folder (when downloaded from GitHub).
-   **Outputs** of GitHub are written to a new local folder named `output_rerun/`

## Setup

```{r setup}
rm(list = ls())

# ---- packages ----
suppressPackageStartupMessages({
  # Data manipulation & plotting
  library(tidyverse)   # dplyr, ggplot2, etc. (%>%, mutate, ggplot, ...)
  library(readxl)      # read_excel()
  
  # Special geoms
  library(ggdist)      # stat_halfeye(), stat_dots()
  library(ggbeeswarm)  # position_beeswarm()
  library(ggpubr)      # theme_pubr(), stat_compare_means()
  
  # Mixed models
  #library(lme4)        # lmer()
  #library(lmerTest)    # p-values/df for lmer models
  library(performance)
  library(DHARMa)
  library(performance) # Model diagnostics
  library(DHARMa)      # Residual diagnostics
  library(car)         # Anova tables
  library(knitr)       # Table formatting
  library(kableExtra)  # Table styling
  library(Cairo)       # PDF rendering
  library(cowplot)     # Plot saving
})
```

## Configuration

```{r}
# ---- path configuration ----
# Script lives in /code, data lives in /source_data
data_dir <- file.path("..", "source_data")
out_dir  <- file.path("..", "output")
SAVE_OUTPUTS <- TRUE 

if (!dir.exists(out_dir)) dir.create(out_dir, showWarnings = FALSE)

# Path helpers
dpath <- function(...) file.path(data_dir, ...)
opath <- function(...) file.path(out_dir, ...)

# Load data
h2b_cells.df <- readxl::read_excel(dpath("2_thr_h2b_cells.xlsx")) %>%
  mutate(replicate = as.factor(replicate))

```

```{r config}
# 
# # ---- configuration ----
# pick_pipeline <- "local"  # "local" or "github"
# 
# # ---- my paths (inputs+outputs) ----
file_path <- "/Users/andrea/Library/CloudStorage/GoogleDrive-andrea.mariossi@gmail.com/My Drive/4.paper_archive/2025_thyroid_pathway_&_metamorphosis/04.SciAdv_rev/source_data"
# out_dir   <- file.path(file_path, "output")
# 
# # ---- github repo paths ----
# repo_dir <- here()
# 
# # fallback: check if repo exists in Downloads folder
# if (!dir.exists(file.path(repo_dir, "data"))) {
#   repo_dir_try <- path.expand("~/Downloads/thyroid_hormone_ciona-main")
#   if (dir.exists(file.path(repo_dir_try, "data"))) {
#     repo_dir <- repo_dir_try
#   }
# }
# 
# # ---- switch ----
# 
# if (pick_pipeline == "local") {
#   data_dir <- file_path
#   SAVE_OUTPUTS <- TRUE
#   dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
#   opath <- function(...) file.path(out_dir, ...)
# 
# } else if (pick_pipeline == "github") {
#   data_dir <- file.path(repo_dir, "data")
#   stopifnot(dir.exists(data_dir))
# 
#   SAVE_OUTPUTS <- TRUE
#   out_dir <- file.path(repo_dir, "output")
#   dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
#   opath <- function(...) file.path(out_dir, ...)
# 
# } else {
#   stop("pick_pipeline must be either 'local' or 'github'")
# }
# 
# dpath <- function(...) file.path(data_dir, ...)
# 
# message("Pipeline: ", pick_pipeline)
# message("Data dir (inputs): ", data_dir)
# message("Outputs enabled: ", SAVE_OUTPUTS)
# if (SAVE_OUTPUTS) message("Outputs dir: ", out_dir)

```

## Theme & Style

```{r theme}
# ---- theme & style ----
custom_theme <- function() {
  theme(
    axis.ticks = element_line(linewidth = 0.5),
    axis.ticks.length = unit(0.09, "cm"),
    legend.key.height = unit(20, "pt"),
    axis.line = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank(),
    strip.background = element_blank()
  )
}

theme_set(theme_pubr(base_size = 13, base_family = "Helvetica") + custom_theme())
no_legend <- theme(legend.position = "none")

# ---- color palette ----
condition_colors <- c("Control" = "#1b9e77", "Mutant" = "#d95f02")
raincloud_colors <- c("Control" = "#A20842", "Mutant" = "#E15759")
replicate_shapes <- c(21, 22, 23)  # circle, square, diamond — all fillable

# ---- pretty p-values for kable ----
format_pvalues <- function(df) {
  df %>%
    mutate(across(
      any_of(c("p.value", "p.value.x", "p.value.y", "adj.p.value")),
      ~ case_when(
        . < 0.001 ~ format(., scientific = TRUE, digits = 2),
        TRUE ~ as.character(round(., 4))
      )
    ))
}
```

## Load Data

```{r load_data}
# h2b_file <- dpath("2_thr_h2b_cells.xlsx") 
# stopifnot(file.exists(h2b_file))
# 
# h2b_cells.df <- read_excel(h2b_file) %>%
#   mutate(replicate = as.factor(replicate))

# display first rows
h2b_cells.df %>%
  head() %>%
  kable(caption = "H2B-positive cells data (first 6 rows)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# summary statistics
summary_data <- h2b_cells.df %>%
  group_by(condition) %>%
  dplyr::summarise(
    mean = mean(h2b_cells),
    sd = sd(h2b_cells),
    n = n(),
    se = sd / sqrt(n)
  )

summary_data %>%
  kable(caption = "Summary statistics by condition") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# replicate-level summary
rep_level <- h2b_cells.df %>%
  group_by(condition, replicate) %>%
  dplyr::summarise(mean_cells = mean(h2b_cells), .groups = "drop")

rep_level %>%
  kable(caption = "Mean cells per replicate") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Statistical Analysis

### Linear model

Fit a simple linear regression to compare H2B cell counts between conditions:

$$
y_{ij} = \beta_0 + \beta_1\,\text{Condition}_i + \epsilon_{ij},
$$

where $y_{ij}$ is the H2B cell count for observation (j) in condition (i) and $\epsilon_{ij}$ is the error

In the model (`lm(h2b_cells ~ condition)`), `condition` is treated as a two-level factor and can be written as an indicator variable:

$$
{h2b-cells} = \beta_0 + \beta_1\,I(\text{Mutant}) + \epsilon,
$$

-   Intercept (β0): the mean H2B cell count in the reference condition.
-   Condition effect (β1): the difference in mean H2B cell count between Mutant and the reference condition:

$$
\beta_1 = \mathbb{E}[y \mid \text{Mutant}] - \mathbb{E}[y \mid \text{Reference}].
$$

```{r mixed_model}
#----------------------------------------------------------------------------------
# linear model
#----------------------------------------------------------------------------------
lm_fit<- lm(h2b_cells ~ condition, data = h2b_cells.df)

# get ANOVA results
anova_results <- anova(lm_fit)
anova_results

summary(lm_fit)
# plot(lm_fit)
# High R² = 0.7966 → Condition explains ~80% of variance
# Highly significant: p < 2.2e-16

# extract p-value for condition effect
p_value <- anova_results$`Pr(>F)`[1]
cat(sprintf("\nP-value for condition effect: %.4f\n", p_value))

# wald chi-square test
cat("\nType II Wald Test:\n")
print(car::Anova(lm_fit, type = 2))

# ---- replicate check ----
# Quick check to verify if 'replicate' acts as a significant batch effect
res_aov <- aov(h2b_cells ~ condition + replicate, data = h2b_cells.df)
summary(res_aov) 
# Note: p=0.173 for replicate indicates no significant batch effect, 

```

### **Linear mixed-effects model -**

-   fixed effects (condition)

-   only Random effects (replicate-to-replicate variability) <!-- -->

-   The random intercept for **replicate** has variance **1.473** (SD **1.214**), indicating **very little variation between experiments** (replicates are highly consistent).

-   The **residual** variance is **68.185** (SD **8.257**), showing that **most variability occurs embryo-to-embryo within a replicate**, rather than across experiments.

-   **conditionMutant** Simple Linear Model:\*\*

-   `Intercept:`**= 66.764** → Mean H2B-positive cell count for control (wild type) embryos.

-   `conditionMutant` = -32.795 (p \< 2.2e-16)\*\* → Mutant embryos have **\~32.8 fewer H2B-positive cells** compared to controls.

-   `Wild type embryos`: \~66.8 H2B-positive cells per embryo.

-   `Mutant embryos`: \~34.0 H2B-positive cells per embryo (66.764 - 32.795).

Mutant effect: A highly significant reduction of approximately 33 H2B-positive cells (p \< 0.0001), representing a \~49% decrease relative to wild type.

### Model Diagnostics

```{r}
#| label: model-diagnostics
#| fig-width: 12
#| fig-height: 10

# 1. visual check of the assumptions of categorical linear model
# (Normality, Homoscedasticity, Linearity, and Outliers)
performance::check_model(lm_fit) 

# 2. Specific Numerical Tests
performance::check_normality(lm_fit)           # Do residuals follow a bell curve?
performance::check_heteroscedasticity(lm_fit)  # Is the spread of data even across groups? 
# Note: p=0.048 is borderline but acceptable given DHARMa results
performance::check_outliers(lm_fit)            # Are any embryos "extreme" enough to skew results?

# 3. DHARMa: Simulated residuals 
# Especially helpful for count data to check for "overdispersion"
simulation_res <- simulateResiduals(lm_fit)
plot(simulation_res)

```

## Main Figure: Raincloud Plot

```{r raincloud_plot, fig.width=12, fig.height=7}
#| fig-width: 12
#| fig-height: 7
#| out-width: 100%
# ___________________
# ---- final raincloud plot with replicate shapes and stats ----
th_raincloud_plot <- ggplot(h2b_cells.df, aes(x = condition, y = h2b_cells, fill = condition)) +
  stat_halfeye(adjust = 0.9, width = 0.6, justification = -0.3, .width = 0,
               point_colour = NA, alpha = 0.5) +
  geom_boxplot(width = 0.12, outlier.shape = NA, alpha = 0.8, color = "black",
               position = position_nudge(x = 0.22)) +
  geom_point(aes(shape = replicate,fill = condition),
             position = position_beeswarm(dodge.width = 0.3, cex = 1.5),
             size = 4,  alpha = 0.3) +
  scale_fill_manual(values = raincloud_colors) +
  scale_shape_manual(values = replicate_shapes) +
  labs(x = "", y = "# H2B-positive cells") +
  custom_theme() +
   theme(
    legend.position = c(0.1,0.2),
    legend.box = "horizontal",
    legend.text = element_text(size = 18),  # Adjust legend key labels
    legend.title = element_text(size = 20, face = "plain") # Adjust legend title
  ) +
  guides(fill = "none", condition = "none",
         shape = guide_legend(title = "Replicate")) +
  stat_compare_means(
    comparisons = list(c("Control", "Mutant")),
    symnum.args = list(cutpoints = c(0, 0.0001),symbols = c("***")),
    label.y     = max(h2b_cells.df$h2b_cells)*1.05,
    size        = 8,
    vjust = .3,
    tip.length = 0,
    bracket.size = 1
  ) 

th_raincloud_plot
```

Mutant embryos have \~33 fewer reporter-positive cells per embryo than controls, consistently across the three experiment

## Alternative Visualizations

```{r alternative_plots, fig.width=12, fig.height=7}
#| fig-width: 12
#| fig-height: 7
#| out-width: 100%
#
# ---- basic boxplot with jitter ----
ggplot(h2b_cells.df, aes(x = condition, y = h2b_cells, fill = condition)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  labs(x = "Condition", y = "H2B-positive cell count") +
  custom_theme() +
  theme(legend.box = "horizontal")

# ---- violin + boxplot ----
ggplot(h2b_cells.df, aes(x = condition, y = h2b_cells, fill = condition)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white") +
  scale_fill_manual(values = condition_colors) +
  labs(x = "Condition", y = "H2B-positive cell count") +
  custom_theme() +
  theme(legend.box = "horizontal")

# ---- half-eye plot + boxplot + dots ----
ggplot(h2b_cells.df, aes(x = condition, y = h2b_cells, fill = condition)) +
  stat_halfeye(
    adjust = 0.5, justification = -0.2, .width = 0, point_colour = NA
  ) +
  geom_boxplot(width = 0.12, outlier.shape = NA) +
  stat_dots(side = "left", justification = 1.1, binwidth = 1) +
  scale_fill_manual(values = condition_colors) +
  labs(x = "Condition", y = "H2B-positive cell count") +
  custom_theme()
```

## Save Plots

```{r save_plots}
if (SAVE_OUTPUTS) {
  
  width_in <- 40
  aspect_ratio <- 3/4
  height_in <- width_in * aspect_ratio
  
  width_mm <- 4 * width_in / 25.4
  height_mm <- 4 * height_in / 25.4
  
  # save with CairoPDF
  CairoPDF(
    opath("fig1I_nkx2.1-mut_thr-reporter.pdf"),
    width = width_mm,
    height = height_mm
  )
  print(th_raincloud_plot)
  dev.off()
  
  # save with cowplot
  cowplot::save_plot(
    plot = th_raincloud_plot,
    base_height = 1.9 * 3.71,
    base_asp = 0.9 * 1.618,
    file = opath("fig1I_nkx2.1-mut_thr-reporter_cowplot.pdf")
  )
  
  message("\nPlots saved to: ", out_dir)
}
```

## Session Info

```{r session_info}
sessionInfo()
```
