---
title: "T4 Tail Regression Analysis"
author: "Andrea Mariossi"
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    toc-depth: 5
    code-fold: false
    code-line-numbers: true
    page-layout: article
    grid:
      sidebar-width: 100px
      body-width: 2000px
      margin-width: 100px
      gutter-width: 3.5rem
execute:
  echo: true
  warning: false
  message: true
fig-width: 12
fig-height: 12
fig-align: center
fig-format: png
editor_options:
  chunk_output_type: inline
---

## Experimental Design: T4-Exposure Time-Course

### Overview

**`Overview`**

For each experimental day (replicate_batch), embryos are obtained from a parent set of (5–7 adults). Eggs are pooled and fertilized using a mixed sperm pool from the same parent set. After fertilization, embryos are randomly allocated into dishes assigned
to one of six conditions (Control and Cond_1–Cond_5), which represent different T4-exposure durations.

### Replicate Batches and Conditions

-   **Number of replicate_batch:** 4 (performed on 4 different days)
-   **Conditions per replicate_batch (6 total):** Control, Cond_1, Cond_2, Cond_3, Cond_4, Cond_5
    -   Cond_1–Cond_5 correspond to different T4-exposure durations

### Sample Size and Measurements

-   **Embryos per dish (approx.):** 30–70
-   **Measured n per dish:** varies by dish (e.g., 33, 37, 42, 47, 49, 50, 53, …)
-   **Measurement:** tail length
-   **Measurement frequency:** each embryo is measured once (one tail-length measurement per individual)

## Setup

```{r setup}
rm(list = ls())

# ---- packages ----
suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
  library(ggpubr)
  library(knitr)
  library(kableExtra)
  library(ggdist)
  library(gghalves)
  library(ggforce)
  library(rstatix)
  library(lme4)
  library(lmerTest)
  library(emmeans)
  library(performance)
  library(DHARMa)
  library(Cairo)
  library(cowplot)
})
```

## Data Loading

```{r}
# ---- paths ----
data_dir <- file.path("..", "source_data")
out_dir  <- file.path("..", "output")

# Uncomment for local absolute path:
# data_dir <- "/Users/andrea/Library/CloudStorage/GoogleDrive-andrea.mariossi@gmail.com/My Drive/4.paper_archive/2025_thyroid_pathway_&_metamorphosis/04.SciAdv_rev/source_data/"

SAVE_OUTPUTS <- FALSE
if (!dir.exists(out_dir)) dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# helpers
dpath <- function(...) file.path(data_dir, ...)
opath <- function(...) file.path(out_dir, ...)

# ---- dataset ----
data_file <- "9_T4_percentange_tail_regression.xlsx"
stopifnot(file.exists(dpath(data_file)))
```

## Theme & Style

```{r}
# ---- theme & style ----
custom_theme <- function() {
  theme(
    axis.ticks = element_line(linewidth = 0.5),
    axis.ticks.length = unit(0.09, "cm"),
    legend.key.height = unit(20, "pt"),
    axis.line = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank(),
    strip.background = element_blank()
  )
}

theme_set(theme_pubr(base_size = 13, base_family = "Helvetica") + custom_theme())

# ---- condition order ----
cond_levels <- c("Control","Cond_1","Cond_2","Cond_3","Cond_4","Cond_5")

# ---- color palette ----
color_scheme <- c(
  "Control"  = "#999999",
  "Cond_1"   = "#d9f0a3",
  "Cond_2"   = "#addd8e",
  "Cond_3"   = "#78c679",
  "Cond_4"   = "#41ab5d",
  "Cond_5"   = "#238443"
)

# ---- pretty p-values for kable ----
format_pvalues <- function(df) {
  df %>%
    mutate(across(
      any_of(c("p.value", "p.value.x", "p.value.y", "adj.p.value", "p.adj")),
      ~ case_when(
        . < 0.001 ~ format(., scientific = TRUE, digits = 2),
        TRUE ~ as.character(round(., 4))
      )
    ))
}
```

## Data Preprocessing

```{r}
# ============================================================================
# T4 TAIL REGRESSION ANALYSIS
# Drug-Exposure Time-Course Experiment
# ============================================================================
# EXPERIMENTAL DESIGN:
# - 4 replicate batches (different days, different parent sets)
# - 6 conditions per batch: Control + Cond_1 to Cond_5 (drug exposure times)
# - ~40-60 embryos per dish
# - Unit of replication: DISH (replicate_batch × condition)
# - Total dishes: 24 (4 batches × 6 conditions)
# - Measurement: tail length (% of body) - one per embryo
# ============================================================================

# Replicate Batch 1 (Day 1, adults A):
# ├── Dish 1: Control (n=47 embryos)
# ├── Dish 2: Cond_1 (n=52 embryos)
# ├── Dish 3: Cond_2 (n=49 embryos)
# ├── Dish 4: Cond_3 (n=45 embryos)
# ├── Dish 5: Cond_4 (n=51 embryos)
# └── Dish 6: Cond_5 (n=48 embryos)
# 
# Replicate Batch 2 (Day 2, adults B):
# ├── Dish 2_Control    → embryos 1, 2, 3, ...
# ├── Dish 2_Cond_1     → embryos 1, 2, 3, ...
# └── ...
# 
# Replicate Batch 3 (Day 3, Parents C)
# └── ...
# 
# ReplicateBatch 4 (Day 4, Parents D)
# └── ...
# 
# Total: 4 batches × 6 conditions = 24 dishes
```

```{r}
# ============================================================================
# T4 TAIL REGRESSION ANALYSIS

# ---- load wide ----
T4_tailReg.df <- readxl::read_excel(dpath(data_file))

T4_tailReg.df %>%
  head() %>%
  kable(caption = "T4 tail regression (wide): first 6 rows") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# ---- tidy long ----
T4_tailReg_tidy <- T4_tailReg.df %>%
  pivot_longer(
    cols = everything(),
    names_to = c("condition", "replicate_batch"),
    names_pattern = "^(.*)_rep(\\d+)$",
    values_to = "tail_length",
    values_drop_na = TRUE
  ) %>%
  mutate(
    condition = factor(condition, levels = cond_levels),
    replicate_batch = factor(replicate_batch),
    dish_id = interaction(replicate_batch, condition, drop = TRUE)
  )

# # quick counts
# counts_by_dish <- T4_tailReg_tidy %>%
#   count(condition, replicate_batch, name = "n_fish") %>%
#   arrange(condition, replicate_batch)
# # Summarise by treatment 
# T4_tailReg_tidy |>
#   filter(!is.na(tail_length)) |>
#   summarise(n_individuals = n(), .by = condition)
# # Summarise by replicate_batch
# T4_tailReg_tidy |>
#   filter(!is.na(tail_length)) |>
#   summarise(n_individuals = n(), .by = replicate_batch)
# # totals embryos counts
# T4_tailReg_tidy |>
#   filter(!is.na(tail_length)) |>
#   count(condition, replicate_batch, name = "n_individuals") |>
#   summarise(total_individuals = sum(n_individuals))

# counts_by_dish %>%
#   kable(caption = "Counts per dish (condition × replicate_batch)") %>%
#   kable_styling(bootstrap_options = c("striped", "hover"))

cat("\n=== DATA STRUCTURE ===\n")
cat("Total embryo measurements:", nrow(T4_tailReg_tidy), "\n")
cat("Number of batches:", n_distinct(T4_tailReg_tidy$replicate_batch), "\n")
cat("Number of conditions:", n_distinct(T4_tailReg_tidy$condition), "\n")
cat("Total dishes:", n_distinct(T4_tailReg_tidy$dish_id), "\n\n")
```

## QC Data

```{r}
#| fig-width: 12
#| fig-height: 7
#| out-width: 100%

# ---- basic boxplot x replicate ----
ggplot(T4_tailReg_tidy, aes(x = condition, y = tail_length, fill = replicate_batch)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Tail Length by Condition",
       subtitle = "Colored by replicate batch",
       x = "Condition", 
       y = "Tail Length") +
  custom_theme()


# ---- basic boxplot x condition + jitter ----
ggplot(T4_tailReg_tidy, aes(x = condition, y = tail_length, fill = condition)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.15, alpha = 0.25, size = 1) +
  scale_fill_manual(values = color_scheme) +
  labs(x = "", y = "") +
  custom_theme() +
  guides(fill = "none")

# ---- violin x condition + boxplot ----
ggplot(T4_tailReg_tidy, aes(x = condition, y = tail_length, fill = condition)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.12, fill = "white", outlier.shape = NA) +
  scale_fill_manual(values = color_scheme) +
  labs(x = "", y = "") +
  custom_theme() +
  guides(fill = "none")
```

## Statistical Analysis

```{r}
# ---- sample size summary ----
sample_summary <- T4_tailReg_tidy %>%
  group_by(condition, replicate_batch) %>%
  summarise(n_embryos = n(), .groups = "drop")

sample_summary %>% 
  pivot_wider(names_from = replicate_batch, values_from = n_embryos) %>%
  kable(caption = "Embryos per dish") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

condition_totals <- sample_summary %>%
  group_by(condition) %>%
  summarise(total_embryos = sum(n_embryos))

condition_totals %>%
  kable(caption = "Total embryos per condition") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# ---- aggregate to dish level ----
dish_stats <- T4_tailReg_tidy %>%
  group_by(replicate_batch, condition, dish_id) %>%
  summarise(
    mean_tail = mean(tail_length, na.rm = TRUE),
    sd_tail   = sd(tail_length, na.rm = TRUE),
    n_embryos = n(),
    se_tail   = sd_tail / sqrt(n_embryos),
    .groups = "drop"
  )

dish_stats %>%
  kable(caption = "Dish-level summary (replicate_batch × condition)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# ---- condition means across batches ----
condition_means <- dish_stats %>%
  group_by(condition) %>%
  summarise(
    mean = mean(mean_tail),
    sd   = sd(mean_tail),
    n_batches = n(),
    se   = sd / sqrt(n_batches),
    ci_lower = mean - qt(0.975, n_batches - 1) * se,
    ci_upper = mean + qt(0.975, n_batches - 1) * se,
    .groups = "drop"
  )

condition_means %>%
  kable(caption = "Condition summary across batches (dish means ± 95% CI)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

#### Blocked ANOVA on Dish Means

$$
\text{mean_tail}_{bc} = \beta_0 + \beta_c \cdot \text{Condition}_c + \gamma_b \cdot \text{Batch}_b + \epsilon_{bc}
$$

Where:

-   $\text{mean_tail}_{bc}$ = mean tail length for batch $b$, condition $c$
-   $\beta_0$ = intercept (baseline tail length for Control in reference batch)
-   $\beta_c$ = fixed effect of condition $c$ (treatment effect, $c \in \{\text{Cond\_1, ..., Cond\_5}\}$)
-   $\gamma_b$ = fixed effect of batch $b$ (blocking factor, $b \in \{2, 3, 4\}$)
-   $\epsilon_{bc} \sim \mathcal{N}(0, \sigma^2)$ = residual error

Design:

-   $b \in \{1, 2, 3, 4\}$ (4 replicate batches: different experiments days)
-   $c \in \{\text{Control, Cond\_1, Cond\_2, Cond\_3, Cond\_4, Cond\_5}\}$ (control + 5 T4 windows-exposure conditions)
-   $n = 24$ dishes (experimental units: $4 \text{ batches} \times 6 \text{ conditions}$)

```{r}
# Design: condition = treatment (fixed)
#         replicate_batch = blocking factor (fixed)
# Why: Each batch is a complete independent replicate containing all conditions

# ---- blocked linear model on dish means ----
fit_dish <- lm(mean_tail ~ condition + replicate_batch, data = dish_stats)

# ANOVA table
anova_dish <- anova(fit_dish)
print(anova_dish)

# Contrasts: each condition vs Control (BH adjusted)
emm_dish <- emmeans(fit_dish, ~ condition)
contrasts_dish <- contrast(emm_dish, method = "trt.vs.ctrl", ref = "Control") %>%
                  summary(adjust = "BH", infer = TRUE)
contrasts_dish %>%
  as.data.frame() %>%
  format_pvalues() %>%
  kable(caption = "Contrasts vs Control (dish means; BH adjusted)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

### Model Diagnostics

```{r}
#| fig-width: 12
#| fig-height: 10

# Model diagnostics
cat("\n--- Checking Model Assumptions ---\n")
par(mfrow = c(2, 2))
plot(fit_dish)
par(mfrow = c(1, 1))
performance::check_model(fit_dish)

# DHARMa is most informative for mixed models:
sim_res <- DHARMa::simulateResiduals(fit_dish)
plot(sim_res)
```

-   **Residuals vs Fitted:** Random scatter (good - no pattern)
-   **Q-Q Plot:** Points follow the line well (normality assumption met)
-   **Scale-Location:** Relatively flat (homogeneity of variance OK)
-   **Residuals vs Factor Levels:** Balanced across conditions (good)

**Results:** F(5,15) = 348.8, p \< 0.001 \*\*\*

T4 windows exposure duration has a **massive effect** on tail regression.

**Batch effect:** Not significant (F = 0.73, p = 0.55)

**4 windows** exposure duration has a **massive effect** on tail regression.

**Batch effect:** Not significant (F = 0.73, p = 0.55)

| Condition  | Mean Difference | 95% CI         | Effect          |
|------------|-----------------|----------------|-----------------|
| **Cond_1** | +29.5%          | \[27.1, 31.9\] | Much longer     |
| **Cond_2** | +18.5%          | \[16.0, 20.9\] | Longer          |
| **Cond_3** | +6.1%           | \[3.7, 8.5\]   | Slightly longer |
| **Cond_4** | +15.5%          | \[13.0, 17.9\] | Longer          |
| **Cond_5** | +22.8%          | \[20.4, 25.2\] | Much longer     |

### Linear Mixed Model with Nested Random Effects

$$
\text{tail_length}_{bcd} = \beta_0 + \beta_c \cdot \text{Condition}_c + u_b + u_{bd} + \epsilon_{bcd}
$$

Where:

-   $\text{tail_length}_{bcd}$ = tail length for embryo $d$ in dish within batch $b$, condition $c$
-   $\beta_0$ = fixed intercept (overall mean for Control)
-   $\beta_c$ = fixed effect of condition $c$ (treatment effect, $c \in \{\text{Cond\_1, ..., Cond\_5}\}$)
-   $u_b \sim \mathcal{N}(0, \sigma^2_{\text{batch}})$ = random intercept for batch $b$ (day/parent variation)
-   $u_{bd} \sim \mathcal{N}(0, \sigma^2_{\text{dish}})$ = random intercept for dish within batch (dish-level variation)
-   $\epsilon_{bcd} \sim \mathcal{N}(0, \sigma^2)$ = residual error (embryo-level variation)

**Random Effects Structure:**

$$
(1 | \text{replicate\_batch/dish\_id}) \equiv (1 | \text{replicate\_batch}) + (1 | \text{replicate\_batch:dish\_id})
$$

This accounts for:

1.  **Batch-level variation** ($\sigma^2_{\text{batch}}$): differences between experimental days/parent sets
2.  **Dish-level variation** ($\sigma^2_{\text{dish}}$): differences between dishes within the same batch
3.  **Embryo-level variation** ($\sigma^2$): measurement variation between embryos within the same dish

**Variance Partitioning:**

$$
\text{Var}(\text{tail\_length}) = \sigma^2_{\text{batch}} + \sigma^2_{\text{dish}} + \sigma^2_{\text{residual}}
$$

**Note:** If random effect variances are estimated as zero (singular fit), this indicates minimal variation at those levels, and a simpler blocked ANOVA on dish means may be more appropriate.

```{r}
# ============================================================================
# Alternative approach: account for dish clustering within the same physical dish
# Batch as a fixed block and Dish as a random effect
# Note: This should give similar results but with more detail on variance components
# ============================================================================

# ---- nested mixed model ----
fit_mixed <- lmer(tail_length ~ condition + (1 | replicate_batch/dish_id), 
                  data = T4_tailReg_tidy)

# boundary (singular) fit ==> variation is at the embryo level, treatment effect is SO strong (F = 348) that it overwhelms any dish-to-dish variation. Dishes within the same condition are extremely similar
# That warning means the random‐effect variance for (1 | replicate_batch) is estimated ~0 (i.e., your 4 runs don’t differ much after accounting for condition), or the model can’t reliably estimate it with so few runs.

cat("\n--- Model Summary ---\n")
print(summary(fit_mixed))

cat("\n--- ANOVA ---\n")
print(anova(fit_mixed))

cat("\n--- Variance Components ---\n")
print(VarCorr(fit_mixed))

# Contrasts from mixed model
emm_mixed <- emmeans(fit_mixed, ~ condition)
contrasts_mixed <- summary(
  contrast(emm_mixed, method = "trt.vs.ctrl", ref = "Control"),
  adjust = "BH"
)

contrasts_mixed %>%
  as.data.frame() %>%
  format_pvalues() %>%
  kable(caption = "Contrasts vs Control (fish-level lmer; BH adjusted)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Mixed Model Diagnostics

```{r}
#| fig-width: 12
#| fig-height: 10

performance::check_model(fit_mixed)

# DHARMa residuals
sim_res <- DHARMa::simulateResiduals(fit_mixed, plot = FALSE)
plot(sim_res)
```

## Visualization

### Dish Means per Batch (Spaghetti Plot)

```{r}
#| label: t4-tail-dish-means
#| fig-width: 12
#| fig-height: 7

spaghetti_plot <- ggplot() +
  geom_line(
    data = dish_stats,
    aes(x = condition, y = mean_tail, group = replicate_batch),
    alpha = 0.4, linewidth = 0.8
  ) +
  geom_point(
    data = dish_stats,
    aes(x = condition, y = mean_tail, shape = replicate_batch),
    size = 3, alpha = 0.7
  ) +
  geom_errorbar(
    data = condition_means,
    aes(x = condition, ymin = ci_lower, ymax = ci_upper),
    width = 0.15, linewidth = 1
  ) +
  geom_line(
    data = condition_means,
    aes(x = condition, y = mean, group = 1),
    linewidth = 1.5
  ) +
  geom_point(
    data = condition_means,
    aes(x = condition, y = mean),
    size = 4.5, shape = 21, fill = "white", stroke = 1.5
  ) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "#999999", linewidth = 1) +
  labs(
    x = "Drug Exposure Duration",
    y = "Mean Tail Length (% of body)",
    shape = "Batch",
    title = "Dish means per batch + overall mean ± 95% CI"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.position = "right"
  )

print(spaghetti_plot)


line_plot <- ggplot() +
  geom_line(
    data = dish_stats,
    aes(x = condition, y = mean_tail, group = replicate_batch),
    alpha = 0.5
  ) +
  geom_point(
    data = dish_stats,
    aes(x = condition, y = mean_tail),
    size = 2.3,
    alpha = 0.8
  ) +
  geom_errorbar(
    data = condition_means,
    aes(x = condition, ymin = ci_l, ymax = ci_u),
    width = 0.15,
    linewidth = 0.8
  ) +
  geom_line(
    data = condition_means,
    aes(x = condition, y = mean, group = 1),
    linewidth = 1.2
  ) +
  geom_point(
    data = condition_means,
    aes(x = condition, y = mean),
    size = 3.6,
    shape = 21,
    fill = "white",
    stroke = 1
  ) +
  labs(
    x = "Condition (T4 exposure time)",
    y = "Mean tail length (dish mean)",
    title = "Dish means per batch (spaghetti) + overall mean ± 95% CI"
  ) +
  custom_theme() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

# Spaghetti plot: dish means per batch + overall mean
spaghetti_plot <- ggplot() +
  # Batch-specific dish means
  geom_line(
    data = dish_stats,
    aes(x = condition, y = mean_tail, group = replicate_batch),
    alpha = 0.4, linewidth = 0.8
  ) +
  geom_point(
    data = dish_stats,
    aes(x = condition, y = mean_tail, shape = replicate_batch),
    size = 3, alpha = 0.7
  ) +
  # Overall mean ± 95% CI
  geom_errorbar(
    data = condition_means,
    aes(x = condition, ymin = ci_lower, ymax = ci_upper),
    width = 0.15, linewidth = 1
  ) +
  geom_line(
    data = condition_means,
    aes(x = condition, y = mean, group = 1),
    linewidth = 1.5
  ) +
  geom_point(
    data = condition_means,
    aes(x = condition, y = mean),
    size = 4.5, shape = 21, fill = "white", stroke = 1.5
  ) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "#999999", linewidth = 1) +
  labs(
    x = "Drug Exposure Duration",
    y = "Mean Tail Length (% of body)",
    shape = "Batch",
    title = "Dish means per batch + overall mean ± 95% CI"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.position = "right"
  )

```

### Figures 2E: Embryo-Level Distribution (Half-Violin + Sina)

```{r}
#| label: t4-tail-distribution_2
#| fig-width: 12
#| fig-height: 7

# comparisons
comparisons <- list(
  c("Control", "Cond_1"),
  c("Control", "Cond_2"),
  c("Control", "Cond_3"),
  c("Control", "Cond_4"),
  c("Control", "Cond_5")
)

# n labels
n_labels <- condition_totals %>%
  mutate(
    label = paste0("n=", total_embryos),
    y_pos = 37
  )

tr_plot <- ggplot(T4_tailReg_tidy, aes(x = condition, y = tail_length)) +
  gghalves::geom_half_violin(
    aes(fill = condition),
    side = "l",
    adjust = 1.3,
    alpha = 0.7,
    width = 1,
    show.legend = FALSE,
    position = position_nudge(x = -0.15)
  ) +
  gghalves::geom_half_boxplot(
    side = "l",
    alpha = 0.9,
    width = 0.2,
    fill = "white",
    color = "black",
    errorbar.draw = FALSE,
    show.legend = FALSE,
    outlier.shape = NA,
    position = position_nudge(x = -0.15)
  ) +
  ggforce::geom_sina(
    aes(fill = condition, shape = replicate_batch),
    size = 2.5,
    bin_limit = 1,
    maxwidth = 0.3,
    color = "black",
    alpha = 0.5
  ) +
  ggdist::stat_pointinterval(
    position = position_nudge(x = -0.15),
    show.legend = FALSE
  ) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "#999999", linewidth = 1) +
  scale_fill_manual(values = color_scheme) +
  scale_y_continuous(limits = c(35, 150), breaks = c(40, 60, 80, 100), oob = scales::squish) +
  labs(x = "Drug Exposure Duration", y = "Tail Length (% of body)", shape = "Batch") +
  custom_theme() +
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = c(0.05, 0.85),
    legend.background = element_rect(fill = "white", color = "black")
  ) +
  geom_text(
    data = n_labels,
    aes(x = condition, y = y_pos, label = label),
    inherit.aes = FALSE,
    size = 5,
    vjust = 1.5
  ) +
  ggpubr::stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    label = "p.signif",
    hide.ns = FALSE,
    tip.length = 0.01,
    label.y = c(105, 115, 125, 135, 145)
  )

tr_plot
```

```{r}
#| label: t4-tail-distribution
#| fig-width: 12
#| fig-height: 7
#| out-width: 100%

# comparisons (Control vs each Cond)
comparisons <- list(
  c("Control", "Cond_1"),
  c("Control", "Cond_2"),
  c("Control", "Cond_3"),
  c("Control", "Cond_4"),
  c("Control", "Cond_5")
)

# stat test (for brackets; fish-level Wilcoxon as a visual aid)
stat.test <- T4_tailReg_tidy %>%
  rstatix::wilcox_test(tail_length ~ condition, comparisons = comparisons) %>%
  rstatix::adjust_pvalue(method = "BH") %>%
  rstatix::add_significance("p.adj") %>%
  rstatix::add_xy_position(x = "condition")


# n labels (total fish per condition)
n_df <- T4_tailReg_tidy %>%
  count(condition, name = "n_fish") %>%
  mutate(
    label = paste0("n=", n_fish),
    y = max(T4_tailReg_tidy$tail_length, na.rm = TRUE) * 1.02
  )

tr_plot <- ggplot(T4_tailReg_tidy, aes(x = condition, y = tail_length)) +
      # Violin plot

  gghalves::geom_half_violin(
    aes(fill = condition),
    side = "l",
    adjust = 1.3,
    alpha = 0.7,
    width = 1,
    show.legend = FALSE,
    position = position_nudge(x = -0.15)
  ) +
  gghalves::geom_half_boxplot(
    side = "l",
    alpha = 0.9,
    width = 0.2,
    fill = "white",
    color = "black",
    errorbar.draw = FALSE,
    show.legend = FALSE,
    outlier.shape = NA,
    position = position_nudge(x = -0.15)
  ) +
  ggforce::geom_sina(
    aes(fill = condition, shape = replicate_batch),
    size = 3,
    bin_limit = 1,
    maxwidth = 0.3,
    color = "black",
    alpha = 0.35
  ) +
  ggdist::stat_pointinterval(
    position = position_nudge(x = -0.15),
    show.legend = FALSE
  ) +
  scale_fill_manual(values = color_scheme) +
  labs(x = "", y = "") +
  custom_theme() +
  theme(
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16)
  ) +
  geom_text(
    data = n_df,
    aes(x = condition, y = y, label = label),
    inherit.aes = FALSE,
    size = 5,
    vjust = 0
  ) +
  ggpubr::stat_pvalue_manual(
    stat.test %>% mutate(p.signif = case_when(
      p.adj < 0.001 ~ "***",
      p.adj < 0.01  ~ "**",
      p.adj < 0.05  ~ "*",
      TRUE ~ "ns"
    )),
    label = "p.signif",
    tip.length = 0.01,
    hide.ns = FALSE
  )

tr_plot
tr_plot2 <- ggplot(T4_tailReg_tidy, aes(x = fct_inorder(condition), y = tail_length)) +
    # Violin plot
    gghalves::geom_half_violin(
      aes(fill = condition),
      side = "l",
      adjust = 1.3,
      alpha = 0.7,
      width = 1,
      show.legend = FALSE,
      position = position_nudge(x = -0.15)
    ) +
    # Boxplot
    gghalves::geom_half_boxplot(
      aes(fill = condition),
      side = "l",
      alpha = 0.9,
      width = 0.2,
      fill = "white",
      color = "black",
      errorbar.draw = FALSE,
      show.legend = FALSE,
      outlier.shape = NA,
      position = position_nudge(x = -0.15)
    ) +
    # Sina plot (individual points)
    ggforce::geom_sina(
      aes(shape = replicate_batch, fill = condition),
      size = 3,
      bin_limit = 1,
      maxwidth = 0.3,
      color = "black"
    # fill = "white"
    # position = position_nudge(x = 0.01)  # Push slightly right
    ) +
    # Point interval
    stat_pointinterval(
      position = position_nudge(x = -0.15),
      show.legend = FALSE
    ) +
    geom_hline(yintercept = 50, linetype = "dashed", color = "#999999", size = 1) +
    # Scales and aesthetics
    scale_y_continuous(limits = c(35, 100), oob = scales::squish) +
    scale_fill_manual(values = color_scheme) +
    labs(x = "", y = "") +
    guides(fill = "none") +
    custom_theme() +
    theme(
      # aspect.ratio = 0.43,
    # coord_fixed(ratio = 0.67 / 2.4),  # Maintain the same height/width ratio
      panel.grid.major.x = element_blank(),
      axis.text.x = element_text(size = 18),
      axis.text.y = element_text(size = 18),
      plot.margin = margin(0, 0, 0, 0, "in"),
      legend.position = c(0.05, 0.5),       
      legend.background = element_rect(fill = "white", color = "black"),
      legend.key = element_blank()
    ) +
  geom_text(
    data = n_df,
    aes(x = condition, y = y, label = label),
    inherit.aes = FALSE,
    size = 9,
    vjust = 1,
    position = position_nudge(x = -0.05)  # align with the half-violin/box
  )


tr_plot2 +
  scale_y_continuous(limits = c(35, 150),breaks = c(40, 60, 80, 100),  oob = scales::squish
  ) +
  ggpubr::stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    label = "p.signif",
    hide.ns = FALSE,
    tip.length = 0.01,
    label.y = c(105, 115, 125, 135, 145)
  )

# Create n= labels for plot
n_labels <- condition_totals %>%
  mutate(label = paste0("n=", total_embryos),
         y_pos = 37)

# Main plot: half-violin + boxplot + individual points
main_plot <- ggplot(T4_tailReg_tidy, aes(x = condition, y = tail_length)) +
  # Half violin
  geom_half_violin(
    aes(fill = condition),
    side = "l",
    adjust = 1.3,
    alpha = 0.7,
    width = 1,
    show.legend = FALSE,
    position = position_nudge(x = -0.15)
  ) +
  # Half boxplot
  geom_half_boxplot(
    side = "l",
    alpha = 0.9,
    width = 0.2,
    fill = "white",
    color = "black",
    errorbar.draw = FALSE,
    show.legend = FALSE,
    outlier.shape = NA,
    position = position_nudge(x = -0.15)
  ) +
  # Individual points (colored by batch)
  geom_sina(
    aes(shape = replicate_batch, fill = condition),
    size = 3,
    bin_limit = 1,
    maxwidth = 0.3,
    color = "black"
  ) +
  # Mean + CI
  stat_pointinterval(
    position = position_nudge(x = -0.15),
    show.legend = FALSE
  ) +
  # Reference line at 50%
  geom_hline(yintercept = 50, linetype = "dashed", color = "#999999", linewidth = 1) +
  # Sample size labels
  geom_text(
    data = n_labels,
    aes(x = condition, y = y_pos, label = label),
    inherit.aes = FALSE,
    size = 5,
    vjust = 1,
    position = position_nudge(x = -0.05)
  ) +
  scale_y_continuous(limits = c(35, 100), oob = scales::squish) +
  scale_fill_manual(values = color_scheme) +
  labs(x = "Drug Exposure Duration", 
       y = "Tail Length (% of body)",
       shape = "Batch") +
  guides(fill = "none") +
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = c(0.05, 0.85),
    legend.background = element_rect(fill = "white", color = "black"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

# Version with significance brackets
plot_with_sig <- main_plot +
  scale_y_continuous(limits = c(35, 150), breaks = c(40, 60, 80, 100), oob = scales::squish) +
  stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    label = "p.signif",
    hide.ns = FALSE,
    tip.length = 0.01,
    label.y = c(105, 115, 125, 135, 145)
  )
```

## Save Outputs

```{r}
if (SAVE_OUTPUTS) {
  # ---- figures ----
  ggsave(opath("T4_tail_distribution.svg"), tr_plot, width = 12, height = 7)
  ggsave(opath("T4_tail_dish_means_spaghetti.svg"), spaghetti_plot, width = 12, height = 7)

  CairoPDF(opath("T4_tail_distribution.pdf"), width = 12, height = 7)
  print(tr_plot)
  dev.off()

  CairoPDF(opath("T4_tail_dish_means_spaghetti.pdf"), width = 12, height = 7)
  print(spaghetti_plot)
  dev.off()

  # ---- tables ----
  readr::write_csv(dish_stats, opath("T4_tail_dish_stats.csv"))
  readr::write_csv(condition_means, opath("T4_tail_condition_means.csv"))
  readr::write_csv(as.data.frame(contrasts_dish), opath("T4_tail_contrasts_vs_control_dish.csv"))
  readr::write_csv(as.data.frame(contrasts_mixed), opath("T4_tail_contrasts_vs_control_fish_lmer.csv"))

  message("\nOutputs saved to: ", out_dir)
}
```

## Session Info

```{r}
sessionInfo()
```
