---
title: "Perchlorate and T4 rescue"
author: "Andrea Mariossi"
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    toc-depth: 5
    code-fold: false
    code-line-numbers: true
    page-layout: article
    grid:
      sidebar-width: 100px
      body-width: 2000px
      margin-width: 100px
      gutter-width: 3.5rem
execute:
  echo: true
  warning: false
  message: true
fig-width: 18
fig-height: 10
fig-align: center
fig-format: png
out-width: "90%"
editor_options:
  chunk_output_type: inline
---

**Normal metamorphosis progression:** - Swimming → Attach (lv_noTR) → Tail regression → Body rotation (jv_wBodyRotation)

Developmental stages/phenotypes rather than tail regression bins:

-   `jv_wBodyRotation`: Juveniles with body rotation (normal endpoint for control at this stage)
-   `tailed_jv`: Tailed juveniles (abnormal - drug-induced defect)
-   `lv_noTR`: Larva attached after swimming but no tail regression
-   `swimming`: Swimming larvae

## Setup

```{r setup, message=FALSE, warning=FALSE}
rm(list = ls())

# ---- packages ----
suppressPackageStartupMessages({library(tidyverse)
library(rstatix)
library(ggpubr)
library(knitr)
library(kableExtra)
library(scales)
library(patchwork)
library(emmeans)
})

```

## Configuration

```{r}
# ---- path configuration ----
# Script lives in /code, data lives in /source_data
data_dir <- file.path("..", "source_data")
out_dir  <- file.path("..", "output")
SAVE_OUTPUTS <- TRUE 

if (!dir.exists(out_dir)) dir.create(out_dir, showWarnings = FALSE)

# Path helpers
dpath <- function(...) file.path(data_dir, ...)
opath <- function(...) file.path(out_dir, ...)

```

## Theme & Style

```{r theme}

# ---- theme & style ----
custom_theme <- function() {
  theme(
    axis.ticks = element_line(linewidth = 0.5),
    axis.ticks.length = unit(0.09, "cm"),
    legend.key.height = unit(20, "pt"),
    axis.line = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank(),
    strip.background = element_blank()
  )
}

theme_set(theme_pubr(base_size = 13, base_family = "Helvetica") + custom_theme())
no_legend <- theme(legend.position = "none")

# ---- color palette ----

# Color palette for developmental stages
stage_colors <- c(
  "jv_wBodyRotation" = "#8f817e",  
  "tailed_jv"        = "#ded4c8", 
  "lv_noTR"          = "#c49f9a",  
  "swimming"         = "#e4c2c9"  
)

# Condition colors
condition_colors <- c(
  "control"        = "#999999",
  "perchlorate"    = "#66c2a5",
  "perchlorate_T4" = "#8da0cb"
)

# ---- pretty p-values for kable ----
format_pvalues <- function(df) {
  df %>%
    mutate(across(
      any_of(c("p.value", "p.value.x", "p.value.y", "adj.p.value")),
      ~ case_when(
        . < 0.001 ~ format(., scientific = TRUE, digits = 2),
        TRUE ~ as.character(round(., 4))
      )
    ))
}
```

## Data Loading & tidy

```{r data-prep}

# Load data
stage_data <- readxl::read_excel(dpath("11_inhibitors_metamorphosis_rescue.xlsx")) |> 
  mutate(
    condition = factor(condition, levels = c("control", "perchlorate", "perchlorate_T4")),
    bioRep = factor(bioRep)
  )
# stage_data <- tibble(
#   condition = c(rep("control", 6), rep("perchlorate", 6), rep("perchlorate_T4", 6)),
#   bioRep = rep(1:6, 3),
#   jv_wBodyRotation = c(43, 58, 55, 61, 39, 66,  # control
#                        29, 35, 16, 14, 23, 30,  # perchlorate
#                        41, 33, 26, 23, 25, 34), # perchlorate_T4
#   tailed_jv = c(1, 0, 0, 0, 0, 0,
#                 5, 6, 13, 6, 9, 25,
#                 4, 2, 0, 1, 11, 7),
#   lv_noTR = c(6, 5, 2, 4, 1, 0,
#               2, 5, 24, 4, 15, 19,
#               14, 6, 9, 7, 16, 2),
#   swimming = c(8, 5, 2, 8, 1, 8,
#                14, 18, 21, 13, 22, 9,
#                9, 5, 4, 13, 4, 8)
# ) %>%
  # mutate(
  #   condition = factor(condition, levels = c("control", "perchlorate", "perchlorate_T4")),
  #   bioRep = factor(bioRep)
  # )

# Convert to long format
stage_long <- stage_data %>%
  pivot_longer(cols = c(jv_wBodyRotation, tailed_jv, lv_noTR, swimming), names_to = "stage", values_to = "n") %>%
  mutate(stage = factor(stage, levels = c("swimming", "lv_noTR", "tailed_jv", "jv_wBodyRotation")))

# Calculate percentages
stage_percent <- stage_long %>%
  group_by(condition, bioRep) %>%
  mutate(total = sum(n),
         percent = n / total * 100) %>%
  ungroup()

# Display data structure
cat("=== DATA STRUCTURE ===\n")
cat("Conditions:", paste(levels(stage_long$condition), collapse = ", "), "\n")
cat("Developmental stages:", paste(levels(stage_long$stage), collapse = ", "), "\n")
cat("Number of biological replicates per condition:", n_distinct(stage_long$bioRep), "\n\n")

# Preview
stage_long %>%
  head(12) %>%
  kable(caption = "Stage data (first 12 rows)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

# Summary Statistics

```{r summary-stats}
summary_stats <- stage_percent %>%
  group_by(condition, stage) %>%
  dplyr::summarise(
    n_reps = n(),
    mean_pct = mean(percent),
    sd_pct = sd(percent),
    se_pct = sd_pct / sqrt(n_reps),
    .groups = "drop"
  )

summary_stats %>%
  kable(caption = "Summary statistics: mean % ± SE per condition and stage", digits = 1) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

```{r biological-metrics}
# Calculate biologically meaningful metrics
biological_summary <- summary_stats %>%
  select(condition, stage, mean_pct) %>%
  pivot_wider(names_from = stage, values_from = mean_pct) %>%
  mutate(
    # Normal complete metamorphosis (most advanced)
    pct_normal_complete = jv_wBodyRotation,
    
    # Arrested at early stages (swimming + lv_noTR)
    pct_arrested = swimming + lv_noTR,
    
    # Drug-induced malformation during tail regression
    pct_malformed = tailed_jv,
    
    # Pre-metamorphic (not yet attached)
    pct_premetamorphic = swimming,
    
    # Attached but incomplete (lv_noTR + tailed_jv)
    pct_attached_incomplete = lv_noTR + tailed_jv
  ) %>%
  select(condition, pct_normal_complete, pct_arrested, pct_malformed, 
         pct_premetamorphic, pct_attached_incomplete)

biological_summary %>%
  mutate(across(where(is.numeric), ~round(., 1))) %>%
  kable(caption = "Biological interpretation of phenotypes", digits = 1) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

cat("\n=== KEY DEVELOPMENTAL OUTCOMES ===\n")
cat("Control: ", round(biological_summary$pct_normal_complete[1], 1), 
    "% complete metamorphosis (normal endpoint)\n")
cat("Perchlorate: ", round(biological_summary$pct_normal_complete[2], 1), 
    "% complete (", 
    round(biological_summary$pct_normal_complete[1] - biological_summary$pct_normal_complete[2], 1), 
    "% reduction)\n")
cat("Perchlorate+T4: ", round(biological_summary$pct_normal_complete[3], 1), 
    "% complete (", 
    round(biological_summary$pct_normal_complete[3] - biological_summary$pct_normal_complete[2], 1), 
    "% improvement)\n\n")

cat("ARREST AT EARLY STAGES:\n")
cat("Control: ", round(biological_summary$pct_arrested[1], 1), "%\n")
cat("Perchlorate: ", round(biological_summary$pct_arrested[2], 1), "% (+", 
    round(biological_summary$pct_arrested[2] - biological_summary$pct_arrested[1], 1), "%)\n")
cat("Perchlorate+T4: ", round(biological_summary$pct_arrested[3], 1), "% (", 
    round(biological_summary$pct_arrested[3] - biological_summary$pct_arrested[2], 1), "%)\n\n")

cat("MALFORMATIONS (tailed_jv):\n")
cat("Control: ", round(biological_summary$pct_malformed[1], 1), "% (baseline)\n")
cat("Perchlorate: ", round(biological_summary$pct_malformed[2], 1), "% (drug-induced)\n")
cat("Perchlorate+T4: ", round(biological_summary$pct_malformed[3], 1), "%\n")
```

# Statistical Analysis: Chi-Square Tests

The data is compositional (sum to total animals per replicate), statistical approach: **Chi-square test**

Does perchlorate disrupt development? (yes from previus figure S10A) BUT

Does T4 rescue the effect? (partial return to normal distribution)

```{r stat_overview}
# Three pairwise chi-squares tell the whole story

# 1. Does perchlorate differ from control?
# ├─ Test: Control vs Perchlorate chi-square
#    → YES (p < 0.05) means there's an effect

# 2.Does T4 reverse the perchlorate effect?
# ├─ Test: Perchlorate vs Perchlorate+T4 chi-square  
#    → YES (p < 0.05) means T4 has a rescue effect

# 3. Is the rescue complete?
# ├─ Test: Control vs Perchlorate+T4 chi-square
#    → NO (p > 0.05) = COMPLETE rescue (T4 restores to control level)
#    → YES (p < 0.05) = PARTIAL rescue  (T4 helps but not fully)

```

## Overall Test

**Question:** Do the treatments alter the distribution of developmental stages?

```{r chisq-overall}
# contingency table
count_table <- stage_long %>%
  group_by(condition, stage) %>%
  summarise(total_n = sum(n), .groups = "drop") %>%
  pivot_wider(names_from = stage, values_from = total_n, values_fill = 0)

count_table %>%
  kable(caption = "Contingency table: condition × stage") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Convert to matrix
count_matrix <- count_table %>%
  # select(-condition) %>%
  column_to_rownames("condition") %>%
  as.matrix()
# rownames(count_matrix) <- count_table$condition

# Chi-square test
chisq_result <- chisq.test(count_matrix)
cat("\n=== Chi-Square Test (Overall) ===\n")
print(chisq_result)

# Effect size (Cramér's V)
V_overall <- sqrt(as.numeric(chisq_result$statistic) /
                  (sum(count_matrix) * min(nrow(count_matrix) - 1, ncol(count_matrix) - 1)))
cat("\nCramér's V:", round(V_overall, 3), "\n")

# Standardized residuals
stdres_overall <- as.data.frame(as.table(chisq_result$stdres)) %>%
  rename(condition = Var1, stage = Var2, stdres = Freq) %>%
  arrange(desc(abs(stdres)))

stdres_overall %>%
  kable(caption = "Standardized residuals (cells driving the effect)", digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Standardized residuals
stdres_overall <- as.data.frame(as.table(chisq_result$stdres)) %>%
  rename(condition = Var1, stage = Var2, stdres = Freq) %>%
  arrange(desc(abs(stdres))) %>%
  mutate(
    interpretation = case_when(
      condition == "control" & stage == "jv_wBodyRotation" & stdres > 0 ~ 
        "Control has MORE normal metamorphosis (expected)",
      condition == "perchlorate" & stage == "jv_wBodyRotation" & stdres < 0 ~ 
        "Perchlorate has LESS normal metamorphosis",
      condition == "perchlorate" & stage == "swimming" & stdres > 0 ~ 
        "Perchlorate has MORE pre-metamorphic arrest",
      condition == "perchlorate" & stage == "tailed_jv" & stdres > 0 ~ 
        "Perchlorate has MORE malformations",
      abs(stdres) > 2 & stdres > 0 ~ "More than expected",
      abs(stdres) > 2 & stdres < 0 ~ "Less than expected",
      TRUE ~ "Near expected"
    )
  )

stdres_overall %>%
  kable(caption = "Standardized residuals: which stages drive differences", digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))


# Calculate key metrics
descriptive_summary <- as.data.frame(count_matrix) %>%
  tibble::rownames_to_column("condition") %>%
  mutate(
    n = rowSums(across(where(is.numeric))),
    # Proportion in most developed stage
    swimming_prop = swimming / n,
    # Proportion in least developed stage
    rotation_prop = jv_wBodyRotation / n,
    # Ordinal score (higher = more developed)
    # Assign scores: rotation=1, tailed=2, lv_noTR=3, swimming=4
    dev_score = (jv_wBodyRotation * 1 + tailed_jv * 2 + lv_noTR * 3 + swimming * 4) / n
  ) %>%
  select(condition, n, rotation_prop, swimming_prop, dev_score) %>%
  arrange(desc(dev_score))

descriptive_summary %>%
  mutate(
    rotation_prop = percent(rotation_prop, accuracy = 0.1),
    swimming_prop = percent(swimming_prop, accuracy = 0.1),
    dev_score = round(dev_score, 2)
  ) %>%
  kable(caption = "Descriptive summary: key developmental metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Residual Heatmap

```{r residual-heatmap, fig.width=10, fig.height=5}
ggplot(stdres_overall, aes(x = stage, y = condition, fill = stdres)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(stdres, 1)), size = 5) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  scale_x_discrete(labels = c("JV w/\nRotation", "Tailed\nJV", "LV\nNo TR", "Swimming")) +
  labs(x = "Developmental Stage",y = "Condition",fill = "Std. Residual") +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
```

## Pairwise Comparisons: T4 Rescue Analysis

```{r pairwise-chisq}
# Pairwise 2×4 chi-square tests
pairs <- combn(rownames(count_matrix), 2, simplify = FALSE)

pairwise_results <- map_dfr(pairs, function(p) {
  sub <- count_matrix[p, , drop = FALSE]
  tst <- chisq_test(sub)
  
  tibble(
    group1 = p[1],
    group2 = p[2],
    statistic = tst$statistic,
    df = tst$df,
    p = tst$p,
    cramer_v = cramer_v(sub)
  )
}) %>%
  mutate(p.adj = p.adjust(p, method = "holm")) %>%
  arrange(p.adj)

# Add interpretations for rescue analysis
rescue_interpretation <- pairwise_results %>%
  mutate(
    test_type = case_when(
      group1 == "control" & group2 == "perchlorate" ~ 
        "1. Perchlorate Effect",
      group1 == "perchlorate" & group2 == "perchlorate_T4" ~ 
        "2. T4 Rescue",
      group1 == "control" & group2 == "perchlorate_T4" ~ 
        "3. Rescue Completeness"
    ),
    result = ifelse(p.adj < 0.05, "Significant", "Not Significant"),
    interpretation = case_when(
      test_type == "1. Perchlorate Effect" & result == "Significant" ~ 
        "Perchlorate significantly disrupts metamorphosis",
      test_type == "2. T4 Rescue" & result == "Significant" ~ 
        "T4 significantly reverses perchlorate-induced disruption",
      test_type == "2. T4 Rescue" & result == "Not Significant" ~ 
        "T4 does NOT rescue perchlorate effects",
      test_type == "3. Rescue Completeness" & result == "Not Significant" ~ 
        "COMPLETE rescue: T4 fully restores to control levels",
      test_type == "3. Rescue Completeness" & result == "Significant" ~ 
        "PARTIAL rescue: T4 improves but doesn't fully restore to control"
    )
  )

rescue_interpretation %>%
  # select(test_type, group1, group2, cramer_v, p.adj, result, interpretation) %>%
  mutate(p = format(p, scientific = TRUE, digits = 2),
         p.adj = format(p.adj, scientific = TRUE, digits = 2) ) %>%
  kable(caption = "T4 Rescue Analysis: Chi-Square Pairwise Test", digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## 

| Condition      | Mean % Swimming | Mean % JV w/ Rotation |
|----------------|-----------------|-----------------------|
| Control        | \~13%           | \~73%                 |
| Perchlorate    | \~33%           | \~25%                 |
| Perchlorate+T4 | \~14%           | \~60%                 |

## QC

### Biological Replicate Variation

```{r qc-facets, fig.width=12, fig.height=8}
ggplot(stage_percent, aes(x = condition, y = percent, fill = stage)) +
  geom_col(position = "stack", color = "black") +
  facet_wrap(~ bioRep, ncol = 3) +
  scale_fill_manual(
    values = stage_colors,
    name = "Developmental Stage",
    labels = c("JV w/ Rotation", "Tailed JV", "LV No TR", "Swimming")
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    x = "Condition",
    y = "% of Animals",
    title = "Developmental Stage Distribution by Biological Replicate"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )
```

## S10B: Stacked Bar Plot with Statistics

```{r main-plot, fig.height=7, fig.width=10}
#| message: false

# # Vertical
# summary_for_plot <- summary_stats %>%
#   group_by(condition) %>%
#   arrange(stage) %>%
#   mutate(
#     ymax_cumul = cumsum(mean_pct),
#     ymin_cumul = lag(ymax_cumul, default = 0),
#     ymid_cumul = (ymin_cumul + ymax_cumul) / 2
#   ) %>%
#   ungroup()
# 
# # Extract p-values for annotations
# pval_annotations <- pairwise_results %>%
#   filter(group1 == "control") %>%
#   mutate(
#     significance = case_when(
#       p.adj < 0.001 ~ "***",
#       p.adj < 0.01 ~ "**",
#       p.adj < 0.05 ~ "*",
#       TRUE ~ "ns"
#     ),
#     condition = factor(group2, levels = c("perchlorate", "perchlorate_T4")),
#     y_pos = 103
#   )

# horizontal
# Set plotting order (bottom -> top after coord_flip)
summary_for_plot <- summary_stats %>%
  mutate(condition = factor(condition, levels = c("perchlorate_T4", "perchlorate", "control"))) %>%
  group_by(condition) %>%
  arrange(stage) %>%
  mutate(
    ymax_cumul = cumsum(mean_pct),
    ymin_cumul = lag(ymax_cumul, default = 0),
    ymid_cumul = (ymin_cumul + ymax_cumul) / 2
  ) %>%
  ungroup()

# Make sure annotations use the same factor levels
pval_annotations <- pairwise_results %>%
  filter(group1 == "control") %>%
  mutate(
    significance = case_when(
      p.adj < 0.001 ~ "***",
      p.adj < 0.01 ~ "**",
      p.adj < 0.05 ~ "*",
      TRUE ~ "ns"
    ),
    condition = factor(group2, levels = c("perchlorate_T4", "perchlorate", "control")),
    y_pos = 103
  )


# Main plot
p_main <- ggplot(summary_for_plot, aes(x = condition, fill = stage)) +
  geom_col(
    aes(y = mean_pct),
    position = position_stack(reverse = TRUE),
    color = "black",
    width = 0.6,
    linewidth = 0.3
  ) +
  geom_errorbar(
    aes(
      ymin = ymin_cumul + mean_pct - se_pct,
      ymax = ymin_cumul + mean_pct + se_pct
    ),
    width = 0.18,
    color = "black",
    linewidth = 0.5
  ) +
  geom_text(
    aes(y = ymid_cumul, label = sprintf("%.1f%%", mean_pct)),
    color = "black",
    size = 4
  ) +
  # Add significance annotations
  geom_text(
    data = pval_annotations,
    aes(x = condition, y = y_pos, label = significance, fill = NULL),
    size = 6,
    fontface = "bold",
    inherit.aes = FALSE
  ) +
  scale_fill_manual(
    values = stage_colors,
    name = "Developmental Stage",
    labels = c("JV w/ Body Rotation", "Tailed JV", "LV No TR", "Swimming")
  ) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    breaks = seq(0, 100, 25),
    limits = c(0, 108),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_x_discrete(labels = c("Control", "Perchlorate", "Perchlorate\n+ T4")) +
  labs(x = "",y = "% individuals") +
  custom_theme()  +
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 12))

 
p_main + 
  coord_flip() +
  scale_x_discrete(labels = c(
  control = "Control",
  perchlorate = "Perchlorate",
  perchlorate_T4 = "Perchlorate\n+ T4"
))

```

### extra stats:

### ANOVA & Contrasts but on each category

```{r anova-normal}

# Consolidated phenotype groupings for ANOVA
phenotype_groups <- stage_percent %>%
  group_by(condition, bioRep) %>%
  summarise(
    normal = percent[stage == "jv_wBodyRotation"],
    malformed = percent[stage == "tailed_jv"],
    arrested = sum(percent[stage %in% c("swimming", "lv_noTR")]),
    .groups = "drop"
  )

summary_stage_percent <- stage_percent %>%
  group_by(condition) %>%
  summarise(
    mean = mean(percent),
    se = sd(percent) / sqrt(n()),
    .groups = "drop"
  )

# Statistical modeling on % Normal Metamorphosi
fit_normal <- lm(normal ~ condition, data = phenotype_groups)
emm_normal <- emmeans(fit_normal, ~ condition)

# Custom contrasts
rescue_contrasts <- list(
  "Perc_Effect" = c(1, -1, 0),  # Control vs Perchlorate
  "T4_Rescue"   = c(0, -1, 1), # Perchlorate vs Perchlorat+T4
  "Completeness" = c(1, 0, -1) # Control vs Perchlorate+T4
)

contrast_results <- contrast(emm_normal, rescue_contrasts, adjust = "none") %>% 
  as.data.frame()

contrast_results  %>%
  mutate(
      sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ "ns"
    ),
    interpretation = case_when(
      contrast == "Perc_Effect" & estimate > 0 & p.value < 0.05 ~ 
        "Perchlorate REDUCES normal metamorphosis",
      contrast == "T4_Rescue" & estimate > 0 & p.value < 0.05 ~ 
        "T4 INCREASES normal metamorphosis vs perchlorate",
      contrast == "Completeness" & p.value > 0.05 ~ 
        "COMPLETE rescue (T4 = Control)",
      contrast == "Completeness" & p.value < 0.05 ~ 
        "PARTIAL rescue (T4 ≠ Control, still reduced)"
    )
  ) %>%
  kable(caption = "Contrasts on % Normal Complete Metamorphosis", digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

### Arrested Development (swimming + lv_noTR)

```{r anova-arrested}
# Calculate combined arrested phenotype
fit_arrested <- lm(arrested ~ condition, data = phenotype_groups)

cat("\n=== ANOVA: % Arrested Development (swimming + lv_noTR) ===\n")
print(anova(fit_arrested))

# Contrasts
emm_arrested <- emmeans(fit_arrested, ~ condition)
contrast_arrested <- as.data.frame(contrast(emm_arrested, rescue_contrasts, adjust = "none"))

contrast_arrested %>%
  mutate(
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ "ns"
    ),
    interpretation = case_when(
      contrast == "Perc_Effect" & estimate < 0 & p.value < 0.05 ~
        "Perchlorate INCREASES developmental arrest",
      contrast == "T4_Rescue" & estimate < 0 & p.value < 0.05 ~
        "T4 REDUCES developmental arrest",
      contrast == "Completeness" & p.value >= 0.05 ~
        "COMPLETE rescue",
      contrast == "Complete Rescue?" & p.value < 0.05 ~
        "PARTIAL rescue"
    )
  ) %>%
  kable(digits = 3, caption = "Contrasts on % Arrested Development") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

## E: Boxplots: Key Phenotypes

```{r}
# comparisons <- list(
#   c("control", "perchlorate"),
#   c("perchlorate", "perchlorate_T4"),
#   c("control", "perchlorate_T4")
# )
# 
# normal_data <- stage_percent %>%
#   filter(stage == "jv_wBodyRotation")
# 
# malformed_data <- stage_percent %>%
#   filter(stage == "tailed_jv")
# 
# arrested_data <- stage_percent %>%
#   filter(stage %in% c("swimming", "lv_noTR")) %>%
#   group_by(condition, bioRep) %>%
#   summarise(percent = sum(percent), .groups = "drop")
# 
# 
# # Normal complete metamorphosis
# p_normal <- ggplot(normal_data, aes(x = condition, y = percent, fill = condition)) +
#   geom_boxplot(outlier.shape = NA, alpha = 0.7) +
#   geom_jitter(width = 0.2, size = 2.5, alpha = 0.6) +
#   scale_fill_manual(values = condition_colors) +
#   scale_x_discrete(labels = c("Control", "Perc", "Perc+T4")) +
#   labs(
#     x = "Treatment",
#     y = "% Normal Complete\nMetamorphosis",
#     title = "Normal Endpoint"
#   ) +
#   stat_compare_means(
#     comparisons = comparisons,
#     method = "t.test",
#     label = "p.signif",
#     size = 3
#   ) +
#   guides(fill = "none") +
#   theme_minimal(base_size = 12)
# 
# # Malformations
# p_malformed <- ggplot(malformed_data, aes(x = condition, y = percent, fill = condition)) +
#   geom_boxplot(outlier.shape = NA, alpha = 0.7) +
#   geom_jitter(width = 0.2, size = 2.5, alpha = 0.6) +
#   scale_fill_manual(values = condition_colors) +
#   scale_x_discrete(labels = c("Control", "Perc", "Perc+T4")) +
#   labs(
#     x = "Treatment",
#     y = "% Malformations\n(Abnormal TR)",
#     title = "Drug-Induced Defects"
#   ) +
#   stat_compare_means(
#     comparisons = comparisons,
#     method = "t.test",
#     label = "p.signif",
#     size = 3
#   ) +
#   guides(fill = "none") +
#   theme_minimal(base_size = 12)
# 
# # Arrested development
# p_arrested <- ggplot(arrested_data, aes(x = condition, y = percent, fill = condition)) +
#   geom_boxplot(outlier.shape = NA, alpha = 0.7) +
#   geom_jitter(width = 0.2, size = 2.5, alpha = 0.6) +
#   scale_fill_manual(values = condition_colors) +
#   scale_x_discrete(labels = c("Control", "Perc", "Perc+T4")) +
#   labs(
#     x = "Treatment",
#     y = "% Arrested\nDevelopment",
#     title = "Metamorphic Arrest"
#   ) +
#   stat_compare_means(
#     comparisons = comparisons,
#     method = "t.test",
#     label = "p.signif",
#     size = 3
#   ) +
#   guides(fill = "none") +
#   theme_minimal(base_size = 12)
# 
# # Combined
# p_boxplots <- p_normal + p_malformed + p_arrested +
#   plot_annotation(
#     title = "T4 Rescue Effects on Metamorphic Phenotypes",
#     theme = theme(plot.title = element_text(size = 14, face = "bold"))
#   )
# 
# p_boxplots


# summary_plot_data <- stage_percent %>%
#   group_by(condition, stage) %>%
#   summarise(mean_pct = mean(percent), se_pct = sd(percent)/sqrt(n()), .groups = "drop") %>%
#   group_by(condition) %>%
#   mutate(cum_pct = cumsum(mean_pct), ymid = cum_pct - (mean_pct/2))
# 

# Faceted Boxplots
p_boxes <- phenotype_groups %>%
  pivot_longer(cols = c(normal, malformed, arrested), names_to = "metric", values_to = "val") %>%
  ggplot(aes(x = condition, y = val, fill = condition)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.1) +
  facet_wrap(~metric, scales = "free_y") +
  scale_fill_manual(values = condition_colors) +
  custom_theme() +
  labs(title = "T4 Rescue Effects on Metamorphic Phenotypes")

 p_boxes
 
 final_table <- phenotype_groups %>%
  group_by(condition) %>%
  summarise(
    `Normal (%)` = paste0(round(mean(normal), 1), " ± ", round(sd(normal)/sqrt(n()), 1)),
    `Malformed (%)` = paste0(round(mean(malformed), 1), " ± ", round(sd(malformed)/sqrt(n()), 1)),
    `Arrested (%)` = paste0(round(mean(arrested), 1), " ± ", round(sd(arrested)/sqrt(n()), 1))
  ) %>%
  mutate(
    `P-value (vs Control)` = c("-", 
                               format.pval(contrast_results$p.value[1], digits = 2), 
                               format.pval(contrast_results$p.value[3], digits = 2))
  )

kable(final_table, align = "lcccc", booktabs = TRUE, 
      caption = "Table 1. Effects of Perchlorate and T4 Rescue on Metamorphic Outcomes (Mean ± SEM)") %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  add_header_above(c(" " = 1, "Phenotype Distribution" = 3, "Stats" = 1))


```

# Interpretation

```{r summary}

cat("\n╔═══════════════════════════════════════════════════════════════╗\n")
cat("║           T4 RESCUE ANALYSIS - FINAL SUMMARY                  ║\n")
cat("╚═══════════════════════════════════════════════════════════════╝\n\n")

cat("PHENOTYPE DISTRIBUTIONS:\n")
cat("───────────────────────────────────────────────────────────────\n")
cat("% Normal Complete Metamorphosis (jv_wBodyRotation):\n")
cat("  Control:         ", sprintf("%5.1f%%", summary_stage_percent$mean_pct[1]), "\n")
cat("  Perchlorate:     ", sprintf("%5.1f%%", summary_stage_percent$mean_pct[2]), 
    " (", sprintf("%+.1f%%", summary_stage_percent$mean_pct[2] - summary_stage_percent$mean_pct[1]), ")\n")
cat("  Perchlorate+T4:  ", sprintf("%5.1f%%", summary_stage_percent$mean_pct[3]), 
    " (", sprintf("%+.1f%%", summary_stage_percent$mean_pct[3] - summary_stage_percent$mean_pct[2]), " vs Perc)\n\n")

cat("CHI-SQUARE PAIRWISE TESTS:\n")
cat("───────────────────────────────────────────────────────────────\n")
perc_effect <- rescue_interpretation %>% filter(test_type == "1. Perchlorate Effect")
t4_rescue <- rescue_interpretation %>% filter(test_type == "2. T4 Rescue")
rescue_complete <- rescue_interpretation %>% filter(test_type == "3. Rescue Completeness")

cat("1. Control vs Perchlorate:\n")
cat("   ", perc_effect$interpretation, "\n")
cat("   p =", format(perc_effect$p.adj, scientific = TRUE, digits = 2), 
    ", Cramér's V =", round(perc_effect$cramer_v, 3), "\n\n")

cat("2. Perchlorate vs Perchlorate+T4:\n")
cat("   ", t4_rescue$interpretation, "\n")
cat("   p =", format(t4_rescue$p.adj, scientific = TRUE, digits = 2), 
    ", Cramér's V =", round(t4_rescue$cramer_v, 3), "\n\n")

cat("3. Control vs Perchlorate+T4:\n")
cat("   ", rescue_complete$interpretation, "\n")
cat("   p =", format(rescue_complete$p.adj, scientific = TRUE, digits = 2), 
    ", Cramér's V =", round(rescue_complete$cramer_v, 3), "\n\n")

# Calculate rescue efficiency
deficit <- summary_stage_percent$mean_pct[1] - summary_stage_percent$mean_pct[2]
recovery <- summary_stage_percent$mean_pct[3] - summary_stage_percent$mean_pct[2]
rescue_pct <- (recovery / deficit) * 100

cat("RESCUE EFFICIENCY:\n")
cat("───────────────────────────────────────────────────────────────\n")
cat("Perchlorate-induced deficit:  ", sprintf("%5.1f%%", deficit), "\n")
cat("T4-mediated recovery:          ", sprintf("%5.1f%%", recovery), "\n")
cat("Rescue efficiency:             ", sprintf("%5.1f%%", rescue_pct), "\n")
cat("(T4 restored ", sprintf("%.1f%%", rescue_pct), 
    " of the perchlorate-induced deficit)\n\n")

cat("CONCLUSION:\n")
cat("───────────────────────────────────────────────────────────────\n")
if(rescue_complete$p.adj > 0.05) {
  cat("✓ T4 COMPLETELY rescues perchlorate-induced metamorphic arrest\n")
  cat("  (Perchlorate+T4 statistically indistinguishable from Control)\n")
} else {
  cat("→ T4 PARTIALLY rescues perchlorate-induced metamorphic arrest\n")
  cat("  (Perchlorate+T4 improved but still differs from Control)\n")
  cat("  Rescue efficiency: ", sprintf("%.1f%%", rescue_pct), "\n")
}

cat("\n")
```

# Session Info

```{r session-info}
sessionInfo()
```
